%%!TEX root = ../../template.tex
\typeout{NT FILE chapter5}

\chapter{System Evaluation}
\label{cha:evaluation}

\prependtographicspath{{Figures/}}

\section{Experimental Evaluation}
\subsection{Setup \& Methodology}
\subsubsection{LUBM Benchmark adaptation}
\subsubsection{Experiments}
\subsection{Experimental Results}
\subsubsection{Repository Size \& (Up)load Time}
\subsubsection{Query Response Time}
\subsubsection{Query Completeness \& Soundness}
\subsubsection{Combined Metric}
\subsubsection{Result Analysis}

\section{Security Analysis}

In this Section we will discuss the security of \gls{SEnTriS}, taking into consideration its security model, defined in Section \ref{ch3:sec:solution-models:security-model}. We will start by enumerating the security properties of the different versions. After

\subsection{Security Properties}
In conformity with the assumptions explained in ..., and based on its definition, we expect all implementations of \gls{SEnTriS} to possess the following \emph{security properties}:

\begin{enumerate}
    \item A user will be able to create triplestores and access tokens, if and only if they have an active session.
    \item A user will be able to interact with a triplestore, if and only if they have a valid access token.
    \item A user will not be able to query triplestores which they have not been given access to.
    \item A user will only be able to execute operations according to its role.
    \item An external observer will not be able to learn any useful information from eavesdropping messages sent between components.
    \item An attacker will not be able to impersonate any component.
    \item An attacker will not be able to tamper with messages sent between components.
\end{enumerate}

Due to the lack of data confidentiality in \gls{SEnTriS}-V0, the following security properties are exclusive to the other two implementations:
\begin{enumerate}
    \item An attacker with access to the \emph{Triplestore} component will learn nothing more than the size of a triplestore and its \emph{search} and \emph{access} patterns.
    \item Regarding \emph{term equality} during intermediate set operations between \emph{bindings}, an attacker with access to the \emph{Proxy} component will learn nothing more than:
        \begin{enumerate}
            \item \emph{SEnTriS V1}: the equality of all terms, belonging to a series of queries, targeting the same triplestore.
            \item \emph{SEnTriS V2} the equality of all terms present in the graph patterns of individual queries. The bindings of different queries are indistinguishable.
        \end{enumerate}
\end{enumerate}

% Plan V0 [[search_0,0, ... search0,n]->[random join_pipeline]^K -> [UNION|OPTIONAL|MINUS|JOIN search0,0, join_0,n, join,k,n]-> DISTINCT | PROJECT | ORDERBY  | SLICE]

\subsection{Leakage Analysis}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}
\subsection{Security Properties Proofs}

%Corolary: sentry v2 leakage is reduced


\section{Complexity Analysis}

\subsection{Storage Complexity}
%Space complexity in terms of RDF terms
\subsubsection{SEnTriS V0}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}

\subsection{Query Evaluation Complexity}
%Time complexity in terms of node search, number of joins and modular exponentiations. Also consider transitivity and expansion depth
\subsubsection{SEnTriS V0}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}

\section{Final Discussion}

All implementations come with a trade-off between data confidentiality, query evaluation leakage and performance.

\gls{SEnTriS} V0 only provides access control, trading data confidentiality for performance.

\gls{SEnTriS} V1 provides both data confidentiality and access control, but suffers from the same leakage problems of \emph{CryptDB}, where an adversary can associate the leakage produced by multiple queries targeting the same triplestore. This facilitates the computation of the frequency distribution of all observed terms, which simplifies the execution of \emph{leakage-abuse} attacks \autocite{Jutla2021-af, Hahn2019-sg, Shafieinejad2021-rb}. This can be especially problematic during \gls{SPARQL} queries evaluation, as they tend to require a high volume of equality checks. A situation further aggravated when we apply query rewriting techniques: the modified queries usually require an even higher number of equality joins.

In \gls{SEnTriS} V2 provides both data confidentiality, access control and reduced leakage. This is achieved at the cost of: 1) an increase in complexity, during set operations between binding tables: 2) an extra communication round during query evaluation, where we have to reverse the masking applied to the \emph{equality tags} and need to fetch, and decrypt, the node values to generate the final results.