%%!TEX root = ../../template.tex
\typeout{NT FILE chapter5}

\chapter{System Evaluation}
\label{cha:evaluation}

\prependtographicspath{{Figures/}}

In this chapter we present the experimental evaluation of \gls{SEnTriS}. We start by analyzing and discussing the solution's storage and query complexity, in Section \ref{ch5:complexity-analysis}. In Section \ref{ch5:setup-methodology}, we present the experimental evaluation conducted: we describe the testing environments; explain the experimental evaluation scenarios; present and discuss the results. Finally, in Section \ref{ch5:security-analysis}, we present \gls{SEnTriS} security properties, analyze the leakage, and discuss the trade-offs between each version of the protocol.

\section{Complexity Analysis}
\label{ch5:complexity-analysis}

In this Section we will be evaluating \gls{SEnTriS} in terms of space complexity for storage of the encrypted data, and time complexity for query evaluation.

\subsection{Storage Complexity}

When evaluating the storage complexity of \gls{SEnTriS}, we need to take into consideration that two different types of encryption operations are done, which generate ciphertexts with different sizes. The size of a ciphertext generated with the XChaCha20-Poly1305 algorithm, depends on the size of the plaintext (plus a constant 16 bytes tag), whereas the size of the ciphertexts generated with the \gls{DGK} cryptosystem depend on the size of the modulus $n$ (usually with 1024 or 2048 bits).

\paragraph{SEnTriS V1 (\emph{Abox})} For every triple parsed, we generate 9 trapdoors (one per keyword combination) and their respective encrypted node. Furthermore, we also generate the trapdoors pointing to the encrypted frequencies of the triple constituents. Finally, for each keyword we generate a trapdoor referencing their encrypted frequencies. 

\paragraph{SEnTriS V2 (\emph{Abox})} For every triple parsed, we generate 9 trapdoors (one per keyword combination), their respective encrypted node and the nodes equality tag, which is also encrypted and stored twice. Furthermore, we generate the trapdoors pointing to the encrypted frequencies of the triple constituents. Finally, for each keyword we generate a trapdoor referencing their encrypted frequencies, and for each node we generate a trapdoor pointing to its encrypted equality tag.

\paragraph{SEnTriS V1 \& V2 (\emph{Tbox})} For every triple, we generate 3 trapdoors referencing each of its nodes, which are also encrypted.

Let $T_a$, $K_a$ and $N_a$ be, respectively, the triples, keywords and nodes of a dataset's \emph{Abox}. Let $T_t$, $K_t$ and $N_t$ be, respectively, the triples, keywords and nodes of the same dataset's \emph{Tbox}. In Table \ref{tbl:storage-complexity:count} we summarize the count of encryption functions executed during an upload (considering no deletions).

%Space complexity in terms of RDF terms

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|}
    \hline
         \textbf{Version} & \textbf{Deterministic} & \textbf{Probabilistic} & \textbf{DGK} \\ \hline
        V1 & $27T_a + K_a + 3T_t$ & $18T_a + Ka + 3T_t$ & 0 \\ \hline
        V2 & $18T_a + K_a + N_a + 3T_t$ & $18T_a + K_a + N_a + 3T_t$ & $18T_a$ \\ \hline
    \end{tabular}
    \caption{Count of encryption operations.}
    \label{tbl:storage-complexity:count}
\end{table}

Now, let $\overline{S}$ be the average size of a ciphertext generated with the XChaCha20-Poly1305 and $\overline{N}$ the size of a ciphertext generated with the \gls{DGK} cryptosystem. In Table \ref{tbl:storage-complexity:formulas} we show the storage complexity for each of the \gls{SEnTriS} versions.

\begin{table}[h!]
    \centering
    \begin{tabular}{|l|l|l|}
    \hline
        \textbf{Version} & \textbf{Storage Formula} & \textbf{Asymptotic Storage} \\
        \hline
        V1 & $(45T_a + 2K_a + 6T_t) * \overline{S}$ & $O\!\left((T_a + K_a + T_t) * \overline{S}\right)$ \\
        \hline
        V2 & $(36T_a + 2K_a + 2N_a + 6T_t) * \overline{S} + 18T_a * \overline{N}$  & $O\!\left((T_a + K_a + N_a + T_t) * \overline{S} + T_a * \overline{N}\right)$ \\
        \hline
    \end{tabular}
    \caption{Storage complexity comparison.}
    \label{tbl:storage-complexity:formulas}
\end{table}


To conclude, in both versions storage complexity scales linearly. Although, in \gls{SEnTriS} V2 the number of ciphertexts increases by approximately $9T_a + 2N_a$ relative to \gls{SEnTriS} V1, the storage overhead is substantially higher due to the increase in size of the ciphertexts generated with the \gls{DGK} cryptosystem.

\subsection{Query Evaluation Complexity}

In this Section, we will analyze the time complexity of \gls{SEnTriS} for \gls{SPARQL} query evaluation. We will do this analysis with and without inference, starting with read queries and then write queries. We will define complexity in terms of query jobs and cryptographic functions executions.

\subsubsection{Read Query Complexity}

 Let $q_S$, $q_J$, $q_M$, $q_O$, $q_P$, $q_{SO}$, $q_D$ be the number of unique query jobs generated from explicitly defined terms, after parsing a \gls{SPARQL} query $q$. Finally, let $D_{\text{CRYPTO}}$ and $E_{\text{CRYPTO}}$ represent a decryption and encryption function using the CRYPTO algorithm.

\paragraph{Query Parsing} During this initial step, we parse the query and generate the query jobs, to include in the query execution plan. Besides the explicitly defined operations in the query body, for each \gls{BGP} pattern, we construct a randomized pipeline of $job_{JOIN}$. We will specify the count of \gls{BGP} patterns as $q_{\text{bgp}}$. Contrary to Jena, the \gls{SEnTriS} protocol does not possess a job abstraction that directly maps to their $op_{\text{BGP}}$, instead we interpret it as a series of $job_{\text{SEARCH}}$ followed by $(n - 1) * job_{\text{JOIN}}$, where $n$ is the number of $job_{\text{SEARCH}}$ (or alternative the number o triple patterns) scoped to the \gls{BGP}. 

During the generation of this pipeline we construct an adjacency matrix to determine the compatibility for join operations, which requires pairwise comparisons between the $job_{\text{SEARCH}}$ generated for the respective \gls{BGP}. We then do breadth first searches, starting from a randomly selected job, until we find a valid path between all jobs. Complexity wise the matrix generation dominates this part of the query parsing, it is quadratic in the number of the \gls{BGP} scoped $job_{\text{SEARCH}}$, but we are more interested in the number of jobs created, which is $(n - 1)$ new $job_{\text{JOIN}}$.

Consequently, the number of jobs generated during query parsing is:

\begin{align}
    \label{equation:query-parsing}
    \begin{split}
        q_{\text{jobs}} = q_S + q_J + q_{\text{bgp}} * (n - 1) + q_M + q_O+ q_P+ q_{SO}+ q_D
    \end{split}
\end{align}

\paragraph{Query Preparation V1} During the query preparation step, for each unique $job_{\text{SEARCH}}$, we generate a trapdoor to retrieve the frequency of its pattern keyword. For each non $\bot$ frequency, we then generate $\overline{\text{f}}$ new trapdoors, where $\overline{\text{f}}$ is the average keyword frequency. As such, we can express the complexity of this phase as:

\begin{align}
    \label{equation:query-preparation-v1}
    \begin{split}
        O_{\text{Prep V1}} &= O \left( q_S * (E_{\text{CHACHA}} + D_{\text{CHACHA}} + \overline{\text{f}} * E_{\text{CHACHA}})\right) \\
        &= O \left( q_S * (D_{\text{CHACHA}} + E_{\text{CHACHA}} * (\overline{\text{f}} + 1))\right)
    \end{split}
\end{align}

\paragraph{Query Preparation V2} Compared to \gls{SEnTriS} V1, each binding sent to the Proxy component is also masked. Let $mask$ be a masking operation, the complexity of this phase is defined as:

\begin{align}
    \label{equation:query-preparation-v2}
    \begin{split}
        O_{\text{Prep V2}} &= O \left( q_S * (E_{\text{CHACHA}} + D_{\text{CHACHA}} + \overline{\text{f}} * E_{\text{CHACHA}} + \overline{\text{f}} * mask)\right) \\
        &= O \left( q_S * (D_{\text{CHACHA}} + E_{\text{CHACHA}} * (\overline{\text{f}} + 1) + \overline{\text{f}} * mask)\right)
    \end{split}
\end{align}

\subsubsection{Query Evaluation Phase Complexity}

The execution of a SPARQL query plan involves sequential processing of a set of jobs, each performing operations such as pattern matching, join, optional join, union, minus, and projection. The overall complexity of the execution can be understood by analyzing the complexity of these individual operations and how they compose according to the query plan.

Let $n$ be the number of jobs in the execution plan, and for each job $J_i$, let its runtime complexity be $T(J_i)$. Then the total execution complexity is:

\[
T_{\text{exec}} = \sum_{i=1}^n T(J_i)
\]

The complexity of each job depends on the operation it performs and the size of its input data.

\begin{table}[h]
\centering
\begin{tabular}{|c|p{5cm}|c|}
\hline
\textbf{Operation} & \textbf{Asymptotic Complexity} \\
\hline
JOIN & $O(m * n * v)$ \\
\hline
OPTIONAL & $O(m * n * v)$ \\
\hline
MINUS & $O(m * n * v)$ \\
\hline
UNION & $O(m + n)$ \\
\hline
PROJECT & $O(v)$ \\
\hline
\end{tabular}
\caption{Complexities of common SPARQL operations. Here $m,n$ are pattern counts, $v$ is number of join variables, $p$ is patterns, and $k$ projected variables.}
\label{tab:sparql-ops}
\end{table}

In practice, the number of join variables $v$ is typically small compared to pattern counts $m, n$, so JOIN and OPTIONAL complexities are roughly quadratic. The main control loop executes each job once, making the total cost linear in the number of jobs but dominated by the complexities of the underlying job operations.

Thus, the overall execution complexity balances a linear traversal through jobs with potentially quadratic or near-linear complexities within each job's operations.

\subsubsection{Inference Complexity Cost}

When we introduce inference capabilities to the query evaluation protocol, we automatically introduce extra complexity from: 1) retrieval of the ontology; 2) query rewriting, which will also affect the total number of query jobs.

\paragraph{Ontology retrieval} The complexity for the retrieval of the ontology depends on the number of trapdoors generated, and values decrypted. 


%TODO: Ontology reconstruction complexity

\paragraph{Query rewriting} The time complexity for the query rewriting algorithm depends heavily on two main factors: the \emph{ontology structure} and the \emph{query content}.

\begin{itemize}
    \item \textbf{Ontology Structure:} The depth and breadth of class and property hierarchies, presence of transitive and inverse properties, and defined class restrictions.
    \item \textbf{Query Content:} The number and type of triple patterns present in the query, and how these relate to the ontology.
\end{itemize}

As it was explained in Section \ref{ch3:query-rewriting}, we use an algorithm that recursively expands the triple patterns present in a \gls{SPARQL} query, up to a parametrized depth.

The complexity of this algorithm is directly related to the various different ways we can expand a query term, based on the ontology's structure, at each expansion step. We denominate this as the \emph{branching factor}. When parsing a class $c$, the rewriting algorithm will consider all subclasses, equivalent classes, intersection operands, and class restrictions related to $c$. Similarly, for a property $p$, expansions will account for subproperties, equivalent properties, inverse and symmetric properties, and any applicable transitive closures.

More formally, given class $c$, let $SC(c)$, $EC(c)$, $IC(c)$ be, respectively, the number of its direct subclasses, equivalent classes, and classes for which $c$ appears as an intersection operand. Let $INT(c)$ be the measure of how many operands are involved in the intersection definitions including $c$, and $R(c)$ be the count of restrictions imposed on $c$. 

Given a property $p$, let $SP(p)$, $EP(p)$ and $INV(p)$, be, respectively, the count of its direct subproperties, equivalent and inverse properties. $SYM(p)$ and $TRANS(p)$ are boolean flags that indicate whether $p$ is symmetric and transitive (1 if yes, 0 otherwise).

Furthermore, let $n$ be the number of triple patterns present in the query, $E$ be the parameter that configures the maximum expansion depth (how many rounds of expansion are allowed for a property $p$ or class $c$) and $T$ be the parameter that configures the maximum transitivity depth (a process requires nested loops).  

With these values, we can specify the branching factors for both the class expansion $B_c$, and the property expansion $B_p$:

\begin{align}
    \label{equation:branching-factors}
    \begin{split}
        B_c &= SC(c) + EC(c) + IC(c) + INT(c) + R(c) \\
        B_p &= SP(p) + EP(p) + SYM(p) + INV(p) + TRANS(p) \times T^2 
    \end{split}
\end{align}

Since there is a maximum depth of expansion, the total complexity for expanding class-related patterns can be defined as $O\left( B_c^E \right)$, and for property-related patterns\footnote{We informally separate $T^2$ from $B_p$ to highlight the quadratic component.} as $O \left( (B_p + T^2)^E) \right)$. The overall time complexity can be defined as:

\begin{align}
    \label{equation:query-rewriting-complexity}
    \begin{split}
        O\left( n \times \max \left( B_c^E,\ (B_p + T^2)^E \right) \right) 
    \end{split}
\end{align}


The asymptotic complexity of the query rewriting algorithm is exponential in terms of maximum expansion depth $E$ and quadratic in terms of transitivity depth $T$.

%TODO: Explain how job number is modified, actually revisit, the join pipeline is only for joining the searches... the order of the other jobs does not change. Number only increased based on branching. Will need to add jobs created to the complexity of the expansion because that is what is interesting

\paragraph{Increase in Query Jobs} If we want to evaluate how does the complexity of the query rewriting algorithm affects the number of jobs generated, we need to analyze which jobs are created and when. 

\begin{table}[h]
    \centering
    \begin{tabular}{|l|l|}
    \hline
    Component & Jobs Create \\
    \hline
    $SC(c)$          & 1 search, 1 union per subclass \\
    $EC(c)$          & 1 search, 1 union per equivalent class \\
    $IC(c)$          & 1 search, 1 union, 1 join per intersection \\
    $INT(c)$         & 1 search, 1 union per intersection operand \\
    $R(c)$           & 1 search, 1 union, 1 minus per restriction \\
    $SP(p)$          & 1 search, 1 union per subproperty \\
    $EP(p)$          & 1 search, 1 union per equivalent property \\
    $SYM(p)$         & 1 search, 1 union if symmetric \\
    $INV(p)$         & 1 search, 1 union if inverse \\
    $TRANS(p) \times T^2$ & 1 search, 1 join, 1 union per transitive step \\
    \hline
    \end{tabular}
    \caption{Jobs Created during pattern expansion}
\end{table}

\vspace{0.5cm}

For a class $c$, the class expansion branching factor $B_c$ combining all job creation is:

\[
B_c = SC(c) \times \left(j_{\text{search}} + j_{\text{union}}\right) + EC(c) \times \left(j_{\text{search}} + j_{\text{union}}\right) + IC(c) \times \left(j_{\text{search}} + j_{\text{union}} + j_{\text{join}}\right) + INT(c) \times \left(j_{\text{search}} + j_{\text{union}}\right) + R(c) \times \left(j_{\text{search}} + j_{\text{union}} + j_{\text{minus}}\right)
\]

For a property $p$, the property expansion branching factor $B_p$ is:

\[
B_p = SP(p) \times \left(j_{\text{search}} + j_{\text{union}}\right) + EP(p) \times \left(j_{\text{search}} + j_{\text{union}}\right) + SYM(p) \times \left(j_{\text{search}} + j_{\text{union}}\right) + INV(p) \times \left(j_{\text{search}} + j_{\text{union}}\right) + TRANS(p) \times T^2 \times \left(j_{\text{search}} + j_{\text{join}} + j_{\text{union}}\right)
\]


The growth of different job types created during query rewriting depends on the ontology's class and property branching factors, $B_c$ and $B_p$, and expansion parameters.

\begin{itemize}
    \item \textbf{Search jobs} grow exponentially with the maximum of the class and property branching factors raised to the expansion depth $E$:
    \item \textbf{Union jobs} follow a similar growth pattern:
    \item \textbf{Join jobs} increase with the branching factors and transitivity depth $T$:
    \item \textbf{Minus jobs} scale linearly with the number of restrictions on class $c$, $R(c)$, and exponentially with expansion depth:
\end{itemize}

\begin{table}[h]
    \centering
    \begin{tabular}{l|c|l}
    \hline
    Job Type    & Growth Formula                        & Notes \\
    \hline
    Search      & $O\left(\max(B_c^E, B_p^E)\right)$           & $B_c, B_p$: class and property branching factors \\
    Union       & $O\left(\max(B_c^E, B_p^E)\right)$           & Mirrors search job growth \\
    Join        & $O\left(\max(B_c^E, B_p^E) \times T\right)$  & $T$: transitivity depth \\
    Minus       & $O\left(R(c) \times \max(B_c^E, B_p^E)\right)$  & $R(c)$: number of class restrictions \\
    \hline
    \end{tabular}
    \caption{Asymptotic growth of job counts by job type with explicit class and property branching factors}
\end{table}



%TODO: Update the overall complexity

\subsubsection{Complexity of write queries}

%TODO: Complexity of delete 
%      Compleity of upload
%      Explain Cases
%          - Generate triples
%          - 



\section{Experimental Evaluation}
\label{ch5:setup-methodology}

\subsection{Setup \& Methodology} 
\subsubsection{LUBM Scenarios}
\subsection{Experimental Results}
\subsubsection{Repository Size \& (Up)load Latency}
\subsubsection{Query Response Latency}
\subsubsection{Query Completeness \& Soundness}
\subsubsection{Result Analysis}

\section{Security Analysis}
\label{ch5:security-analysis}

In this Section we will discuss the security of \gls{SEnTriS}, taking into consideration its security model. We will start by formalizing the security guaranties of the SSE-SPARQL-EVAL protocol, and finish by defining the security properties of \gls{SEnTriS}.

\subsection{Security guaranties of the SSE-SPARQL-EVAL protocol}

We have defined the security model in Section \ref{ch3:sec:solution-models:security-model}, but we will additionally consider that an adversary $\mathcal{A}$ can only compromise either the Triplestore or the Proxy components, not both simultaneously.

\subsection{Leakage Definition}

Let $T(n)$, $E(n)$, $T_{eq}$, and $E(eq)$ be respectively, node trapdoors, encrypted nodes, equality tag trapdoors, and encrypted trapdoors, stored at the triplestore. Let $Bt$, $O(V)$, $Q_r$ be, respectively, the set of binding tables, obfuscated variables and query result, temporarily stored in memory during query evaluation at the Proxy. We will denote the randomized query execution plan as $\pi(Qp)$. 

Furthermore, $n$ is a $k$-bit \gls{RSA} modulus, $m' \in \mathbb{Z}^*_n$ is a masking value and the multiplicative blinding operator is denoted as $B(\cdot)$. The encryption and decryption schemes are specified as $E_{\text{CRYPTO}}(\cdot)$ and $D_{\text{CRYPTO}}(\cdot)$, where $\text{CRYPTO}$ is the cryptographic algorithm. The cryptographic keys are specified as $K_{\text{CRYPTO}}$. 

Let $I(\cdot)$ be a function that, when applied to a trapdoor, reveals its associated dataset. Finally, let $\sigma(\cdot)$ be a function that, when applied to ciphertexts, captures their equivalence and frequency. More specifically, $\sigma(E_{\text{DET}}(\mathrm{bindings}))$ reveals which ciphertexts in the binding tables are equal and how frequently they occur, without disclosing the plaintext, whereas $\sigma(D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags}))))$ represents the equality and frequency of equality tags after applying multiplicative blinding.

We define leakage as a function $\mathcal{L}_{C}^{V}$, where version $V \in \{V1,V2\}$ and component $C \in \{\mathrm{Triplestore}, \mathrm{Proxy}\}$:

\begin{align}
    \label{equation:leakage-functions-triplestores}
    \begin{split}
        \mathcal{L}_{Triplestore}^{V1} &= ( T(n), E(n), I(T(n))) \\
        \mathcal{L}_{Triplestore}^{V2} &= ( T(eq), E(n), E(eq), m, n, I(T(eq)), I(E(eq))) \\
    \end{split}
\end{align}

\begin{align}
    \label{equation:leakage-functions-proxy}
    \begin{split}
        \mathcal{L}_{Proxy}^{V1} &= (Bt, O(V), E_{DET}(\mathrm{bindings}), \sigma(E_{DET}(\mathrm{bindings})), \pi(Qp), Q_r, K_{\text{RND}}) \\
        \mathcal{L}_{Proxy}^{V2} &= (Bt, O(V), D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags}))), \sigma(D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags})))), \\ &\;\;\;\;\;\pi(Qp), Q_r, K_{\text{DGK}_{eq}}) \\
    \end{split}
\end{align}

\subsection{Indistinguishability Games}

For each version $V \in \{V1,V2\}$ and component $C \in \{\mathrm{Triplestore}, \mathrm{Proxy}\}$, the indistinguishability game $G^V_C$ proceeds as follows:

\begin{enumerate}
    \item \textbf{Adversary}: $\mathcal{A}$, restricted to component $C$, chooses two challenge query sets $Q_0$ and $Q_1$.
    \item \textbf{Challenger}: Flips a fair random bit $b \in \{0,1\}$, then runs the SSE-SPARQL-EVAL protocol version $V$ on query set $Q_b$.
    \item \textbf{Challenger}: sends $\mathcal{A}$ the resulting \emph{real view} $\mathrm{RealView}_C^{V}$ seen by component $C$.
    \item \textbf{Simulator}: $\mathcal{S}_C^{V}$, given only leakage $\mathcal{L}_C^{V}(Q_b)$, produces a \emph{simulated view} $\mathrm{SimView}_C^{V}$.
    \item \textbf{Adversary}: $\mathcal{A}$ receives either $\mathrm{RealView}_C^{V}$ or $\mathrm{SimView}_C^{V}$, chosen at random, and attempts to guess $b$.
\end{enumerate}

The adversary's advantage is measured as

\begin{align}
    \label{equation:adversary-advantage}
    \begin{split}
        \mathrm{Adv}_{\mathcal{A}}^{V,C} = \left| \Pr[ \mathcal{A}(\mathrm{RealView}_C^{V})=b ] - \Pr[ \mathcal{A}(\mathrm{SimView}_C^{V})=b ] \right|
    \end{split}
\end{align}


IND-CKA1 or IND-CKA2 holds if the advantage is negligible for any polynomial-time adversary $\mathcal{A}$.


\subsection{Security Proofs (SSE-SPARQL-EVAL V1)}

\begin{theorem} 
    Against an \emph{honest-but-curious} adversary which can compromise either the Triplestore or the Proxy component, the SSE-SPARQL-EVAL V1 protocol achieves \gls{IND-CKA1} security, but not \gls{IND-CKA2} security.
\end{theorem}

\begin{proof} \hfill

    \paragraph{Triplestore Simulator $\mathcal{S}_{Triplestore}^{V1}$}

    \begin{enumerate}
        \item Receives leakage $\mathcal{L}_{Triplestore}^{V1}$.
        \item Generate simulated random trapdoors $T'(n)$ independent of the queries, but consistent with targeted dataset (preserving the dataset association).
        \item Generate simulated randomly encrypted ciphertexts $E'(n)$, referenced by $T'(n)$.
        \item Output $\mathrm{SimView}_{Triplestore}^{V1} = (E'(n), T'(n), I(T(n)))$.
    \end{enumerate}

    \paragraph{Running $G^{V1}_{Triplestore}$}

    \begin{enumerate}
        \item \textbf{Adversary}: $\mathcal{A}$ picks query sets $Q_0, Q_1$.
        \item \textbf{Challenger}: flips $b$, runs protocol SSE-SPARQL-EVAL on $Q_b$ and outputs \linebreak$\mathrm{RealView}^{V1}_{Triplestore}$.
        \item \textbf{Simulator}: $\mathcal{S}_{Triplestore}^{V1}$ receives $\mathcal{L}_{Triplestore}^{V1}$ and outputs $\mathrm{SimView}_{Triplestore}^{V1}$.
        \item \textbf{Adversary}: $\mathcal{A}$ attempts to output $b$. 
    \end{enumerate}

    \paragraph{Proxy Simulator $\mathcal{S}_{Proxy}^{V1}$}

    \begin{enumerate}
        \item Receives leakage $\mathcal{L}_{Proxy}^{V1}$.
        \item Generates a new secret key $K'_{\text{RND}}$.
        \item Generates $E_{DG}(\mathrm{bindings'})$ as follows: for each binding in $E_{DET}(\mathrm{bindings})$, creates a fresh deterministic ciphertext according to $\sigma(E_{DET}(\mathrm{bindings}))$, and keeps track of the mapping between them.
        \item Generates a new set of variables $O(V')$ and constructs a query execution plan, respecting the structure of $\pi(Qp)$. The new plan $\pi(Qp')$ has equivalent jobs, with new random identifiers, that use variables from $O(V')$.
        \item Creates $Bt'$ as follows: for each binding table in $Bt$, creates a new binding table populated with values from $E_{DET}(\mathrm{bindings'})$, according to the saved mappings.
        \item Constructs a simulated $Q_r'$, based on $Q_r$, but filled with values from $E_{DET}(\mathrm{bindings'})$ and according to the saved mappings.
        \item Output $\mathrm{SimView}_{Proxy}^{V1} = (Bt', O(V'), E_{DET}(\mathrm{bindings'}), \sigma(E_{DET}(\mathrm{bindings'})), \pi(Qp'), Q_r', K'_{\text{RND}})$.
    \end{enumerate}

    \paragraph{Running $G^{V1}_{Proxy}$}

    \begin{enumerate}
        \item \textbf{Adversary}: $\mathcal{A}$ picks query sets $Q_0, Q_1$.
        \item \textbf{Challenger}: flips $b$, runs protocol SSE-SPARQL-EVAL on $Q_b$ and outputs \linebreak$\mathrm{RealView}^{V1}_{Proxy}$.
        \item \textbf{Simulator}: $\mathcal{S}_{Proxy}^{V1}$ receives $\mathcal{L}_{Proxy}^{V1}$ and outputs $\mathrm{SimView}_{Proxy}^{V1}$.
        \item \textbf{Adversary}: $\mathcal{A}$ attempts to output $b$. 
    \end{enumerate}

    The semantic security of the cryptographic primitives in use, ensure that \linebreak $\mathrm{Adv}_{\mathcal{A}}^{Triplestore, \; V1}$ is negligible, even after consecutive runs. 
    
    Due to the deterministic encryption of bindings, with a single run of $G^{V1}_{proxy}$, \linebreak $\mathrm{Adv}_{\mathcal{A}}^{Proxy, \; V1}$ is negligible. However, consecutive runs of $G^{V1}_{Proxy}$ will allow $\mathcal{A}$ to keep track of past views, their respective plan size, variable count, ciphertexts and frequency. Because ciphertexts are deterministic, $\mathcal{A}$ will be able to associate ciphertext between different views and construct frequency tables. If they always choose the same set of queries, eventually they will see repeated views. Thus, $\mathrm{Adv}_{\mathcal{A}}^{Triplestore, \; V1}$ stops being negligible.
\end{proof}

\subsection{Security Proofs (SSE-SPARQL-EVAL V2)}

\begin{theorem} 
    Against an \emph{honest-but-curious} adversary which can compromise either the Triplestore or the Proxy component, the SSE-SPARQL-EVAL V2 protocol achieves \gls{IND-CKA1} and \gls{IND-CKA2} security.
\end{theorem}

\begin{proof} \hfill

    \paragraph{Triplestore Simulator $\mathcal{S}_{Triplestore}^{V2}$}

    \begin{enumerate}
        \item Receives leakage $\mathcal{L}_{Triplestore}^{V2}$.
        \item Generate simulated random trapdoors $T'(eq)$ independent of the queries, but consistent with targeted dataset (preserving the dataset association).
        \item Generate simulated randomly encrypted ciphertexts $E'(eq)$, referenced by  $T'(eq)$, and $E'(n)$, referenced by $E'(eq)$.
        \item Samples a new mask $m' \in \mathbb{Z}^*_n$ and computes $B(E'(eq))$.
        \item Output $\mathrm{SimView}_{Triplestore}^{V2} = ( T'(eq), E'(n), E'(eq), I(T'(eq)), I(E'(eq)), m', n, B(E'(eq)))$.
    \end{enumerate}

    \paragraph{Running $G^{V2}_{Triplestore}$}

    \begin{enumerate}
        \item \textbf{Adversary}: $\mathcal{A}$ picks query sets $Q_0, Q_1$.
        \item \textbf{Challenger}: flips $b$, runs protocol SSE-SPARQL-EVAL on $Q_b$ and outputs \linebreak$\mathrm{RealView}^{V2}_{Triplestore}$.
        \item \textbf{Simulator}: $\mathcal{S}_{Triplestore}^{V2}$ receives $\mathcal{L}_{Triplestore}^{V2}$ and outputs $\mathrm{SimView}_{Triplestore}^{V2}$.
        \item \textbf{Adversary}: $\mathcal{A}$ attempts to output $b$. 
    \end{enumerate}

    \paragraph{Proxy Simulator $\mathcal{S}_{Proxy}^{V2}$}

    \begin{enumerate}
        \item Receives leakage $\mathcal{L}_{Proxy}^{V2}$.
        \item Generates a new secret key $K'_{\text{DGK}_{eq}}$.
        \item Generates $E_{DGK_{DET}}(\mathrm{bindings'})$ as follows: for each binding in \linebreak $D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags})))$, creates two values one deterministic and another randomized. These values have similar sizes to, respectively, $D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags})))$ and to $B(E_{\text{DGK}}(\mathrm{eqTags})))$ and respect $\sigma(D_{\text{DGK}_{eq}}(B(E_{\text{DGK}}(\mathrm{eqTags}))))$. The simulator keeps track of the mapping between them. We will refer to the deterministic and random values as $\text{DET}(eq_tags')$ and $\text{RND}(eq_tags')$.
        \item Generates a new set of variables $O(V')$ and constructs a query execution plan, respecting the structure of $\pi(Qp)$. The new plan $\pi(Qp')$ has equivalent jobs, with new random identifiers, that use variables from $O(V')$.
        \item Creates $Bt'$ as follows: for each binding table in $Bt$, creates a new binding table populated with values from $\text{DET}(eqTags')$, according to the saved mappings.
        \item Constructs a simulated $Q_r'$, based on $Q_r$, but filled with values from $\text{RND}(eqTags')$ and according to the saved mappings.
        \item Output $\mathrm{SimView}_{Proxy}^{V2} = (Bt', O(V'), \text{DET}(eqTags'), \sigma(\text{DET}(eqTags')), \pi(Qp'), Q_r',$ \linebreak $K'_{\text{DGK}_{eq}})$.
    \end{enumerate}

    \paragraph{Running $G^{V2}_{Proxy}$}

    \begin{enumerate}
        \item \textbf{Adversary}: $\mathcal{A}$ picks query sets $Q_0, Q_1$.
        \item \textbf{Challenger}: flips $b$, runs protocol SSE-SPARQL-EVAL on $Q_b$ and outputs \linebreak$\mathrm{RealView}^{V2}_{Proxy}$.
        \item \textbf{Simulator}: $\mathcal{S}_{Proxy}^{V2}$ receives $\mathcal{L}_{Proxy}^{V2}$ and outputs $\mathrm{SimView}_{Proxy}^{V2}$.
        \item \textbf{Adversary}: $\mathcal{A}$ attempts to output $b$. 
    \end{enumerate}

    The semantic security of the cryptographic primitives in use, ensure that \linebreak $\mathrm{Adv}_{\mathcal{A}}^{Triplestore, \; V2}$ is negligible, even after consecutive runs. 
    
    Like in version V1, $G^{V2}_{proxy}$, $\mathrm{Adv}_{\mathcal{A}}^{Proxy, \; V1}$ is negligible. On the contrary, consecutive runs of $G^{V2}_{Proxy}$ do not provide any change in $\mathrm{Adv}_{\mathcal{A}}^{Proxy, \; V1}$. The multiplicative blinding ensures that deterministically encrypted equality tags are not linkable between different views.
\end{proof}


\subsection{Security Properties}

From the security analysis of the SSE-SPARQL-EVAL protocol, and taking into account the \gls{RBAC} features of \gls{SEnTriS} we can now define a set of security properties:

\begin{enumerate}
    \item A user will be able to create triplestores and access tokens, if and only if they have an active session.
    \item A user will be able to interact with a triplestore, if and only if they have a valid access token.
    \item A user will not be able to query triplestores which they have not been given access to.
    \item A user will only be able to execute operations according to its role.
    \item An external observer will not be able to learn any useful information from eavesdropping messages sent between components.
    \item An attacker will not be able to impersonate any component.
    \item An attacker will not be able to tamper with messages sent between components.
    \item An attacker with access to the \emph{Triplestore} component will learn nothing more than the size of a triplestore and its \emph{search} and \emph{access} patterns.
    \item Regarding \emph{term equality} during intermediate set operations between \emph{bindings}, an attacker with access to the \emph{Proxy} component will learn nothing more than:
        \begin{enumerate}
            \item \emph{\gls{SEnTriS} V1}: the equality of all terms, belonging to a series of queries, targeting the same triplestore.
            \item \emph{\gls{SEnTriS} V2} the equality of all terms present in the graph patterns of individual queries. The bindings of different queries are indistinguishable.
        \end{enumerate}
\end{enumerate}

Properties 1-4 are derived from the \gls{RBAC} features of \gls{SEnTriS}, properties 5-7 are guaranteed by the secure communication between components, which follow the \glsxtrshort{HTTPS} \autocite{noauthor_undated-yf} protocol and \glsxtrshort{TLS} \autocite{noauthor_undated-kt} data encryption with mutual authentication. Finally, property 8-9 are held by the \gls{IND-CKA1} and \gls{IND-CKA2} security guaranties of the SSE-SPARQL-EVAL protocol.

\section{Final Discussion}

Both implementations come with a trade-off between security guaranties and performance.

\gls{SEnTriS} V1 provides both data confidentiality and access control, but does not achieve \gls{IND-CKA2} security. An adversary can associate the leakage produced by multiple queries targeting the same triplestore. This facilitates the computation of the frequency distribution of all observed terms, which simplifies the execution of \emph{leakage-abuse} attacks \autocite{Jutla2021-af, Hahn2019-sg, Shafieinejad2021-rb}. This can be especially problematic during \gls{SPARQL} query evaluation, as these may require a high volume of equality checks. A situation further aggravated when we apply query rewriting techniques.

\gls{SEnTriS} V2 provides both data confidentiality, access control and \gls{IND-CKA2} security. This is achieved at the cost of an increase in time and space complexity.

From the \gls{LUBM} benchmark results, we argue that both versions remain practical to use. They support a wide range of \gls{SPARQL} queries, provide reasonable upload and query response latency, and achieve perfect results for soundness and completeness.