%%!TEX root = ../../template.tex
\typeout{NT FILE chapter3}

\chapter{SEnTriS: A Searchable Encrypted Triplestore Sharing Protocol}
\label{cha:solution}

\prependtographicspath{{Figures/}}

In this chapter we will present the main contribution of our research. In Section \ref{ch3:sec:solution-models}, we introduce \gls{SEnTriS} and define the system and adversary model. In Section \ref{ch3:sec:solution-definition} we formally define the proposed solution, specify its enforcement of security guaranties and present the \gls{SPARQL} query evaluation pipeline: triple pattern encryption schemes,  high level execution pipelines and protocols, reasoning preservation techniques and list the supported query types.

\section{Solution Models}
\label{ch3:sec:solution-models}

\subsection{Overview}

The current \emph{state-of-the-art} solutions solely focus on the problem of secure \gls{RDF} data outsourcing, leaving the problem of dynamic updates out of scope. Furthermore, these systems do not have reasoning capabilities, suffer from leakage problems and, frequently, solely provide search over simple \emph{triple patterns}.

Considering all these limitations we developed \gls{SEnTriS}, a solution for secure and cooperative development of semantic knowledge bases.

\gls{SEnTriS} possesses role-based \emph{access control} and can execute secure read and write operations. \mytodo{Do not forget to mention verification of authenticity and data integrity as future work} It supports most \gls{SPARQL} operations and retains reasoning capabilities while evaluating queries.

\gls{SEnTriS} is designed for a multi-user setting. It is purely a software solution, not requiring any special hardware, where storage and heavy computations are outsourced to external servers.

\subsection{System Model}
\label{ch3:sec:solution-models:system-model}

\paragraph{Overview} In \gls{SEnTriS}, authenticated \emph{users} are able to create \emph{encrypted triplestores}. Within a triplestore they are able to upload, modify and query semantic data. Triplestores can be shared between users with varying degrees of access.

\gls{SEnTriS} follows the usual \gls{SE} schemes architecture. It is a \emph{client/server} application which outsources data storage and heavy computations to third parties. It follows the \gls{M/M} model with \emph{secrets sharing} between trusted users.

\paragraph{System Architecture} \gls{SEnTriS} is a multi-party system comprised of 5 components:

\begin{itemize}
    \item \emph{Client}: a public facing component where users can authenticate themselves and create, upload, share, request or revoke access to and query triplestores.
    \item \emph{Triplestore}: a storage component where all encrypted data is saved;
    \item \emph{Secure Computation Proxy}: this component executes secure query plans and set operations between encrypted \emph{bindings};
    \item \emph{\gls{IAM} Provider}: a component that handles authentication and validates access policies.
    \item \emph{Vault}: a component to store confidential data required by the cryptographic protocols (e.g. cryptographic keys);
\end{itemize}

Each individual component is a web server that exposes a \glsxtrshort{REST} \gls{API}. All communication channels between components are secure and reliable. They follow the \glsxtrshort{HTTPS} \autocite{noauthor_undated-yf} protocol and \glsxtrshort{TLS} \autocite{noauthor_undated-kt} data encryption with mutual authentication.  Figure \mytodo{Add system architecture figure} depicts a high level architecture of the solution.

\paragraph{Role-Based Access Control} In \emph{SEnTriS} roles grant different permissions to the users, which restrict the type of operations they can execute within the system. Users can have the \emph{ADMIN}, \emph{PRIVILEGED} and \emph{BASIC} roles. For ease of communication from now onwards, we will refer to a user with each role, respectively, as \emph{admin}, \emph{privileged} and \emph{basic} user.

\emph{Basic} users can only query existing triplestores, if they have been given access to do so. They can request \emph{read} or \emph{write} access to existing triplestores.

\emph{Privileged} users have all the permissions of a \emph{basic} user. Additionally, they are able to create new triplestores, grant or revoke read/write access to their triplestores, transfer ownership to other users and process access requests targeting triplestores they own.

\emph{Admin} users have all the permissions of a \emph{privileged} user but can also manage users, any triplestore and process role change requests.


\paragraph{M/M setting} In \gls{SEnTriS} access to triplestores is enforced through the use of \emph{ephemeral access tokens}. These tokens are only obtainable by authenticated users that belong to the respective triplestore's read and write \gls{ACL}.

These tokens ensure that the sharing of \emph{secrets}, needed for the cryptographic protocols, and that the write/read operations are only done by users that have the required permissions to do so.

Correctness of concurrent write and read operations is achieved with a \emph{lock-based} approach when issuing access tokens. \emph{Read access tokens} can be issued at anytime, but only one \emph{write access token}, per triplestore, can exist at any given moment.

\subsection{Security Model}
\label{ch3:sec:solution-models:security-model}

We will consider an \emph{honest-but-curious} adversary. This adversary will try to eavesdrop communication between servers and, if it has access to the external servers where the semantic data and execution logs are stored, try to infer or decrypt sensitive information.

Following the architecture of \gls{SEnTriS}, this means that the \emph{Triplestore} and \emph{Proxy} components will be the targets of attacks. We will consider that the other components and communication channels are trusted, reliable and secure. We will also assume that authenticated users are trusted and non-malicious. Such users can read sensitive information and security parameters of the triplestores they have created or have been given access to.

When targeting these components, the attackers will try to decrypt data and infer sensitive information from the leakage of the protocols in use.

Additionally, we will not consider the case where attacks involve \emph{revoked users}. The role-based access controls prevent these users from interacting (via client) with triplestores they do not have permissions to. Nevertheless, we have not implemented a key rotation, or re-encryption protocol for data. As such, an adversary collaborating with these users, or an attack from such users, will always succeed in the decryption the data of all the triplestores they used to have access to. They can still possess the cryptographic secrets used by the protocols, which makes it trivial to decrypt leaked information. 

\section{Solution Definition}
\label{ch3:sec:solution-definition}

\subsection{Overview}

The functioning of a query evaluation system can be described as a pipeline of sequential operations. Firstly, if the system supports authentication, there is an authentication step. An authenticated user will be able to submit a series of queries while their session is active. A submitted query is parsed, and given the specified heuristics, the system generates a series of query execution plans, which consist in sequences of logical operations (e.g. data retrieval, update or execution of an equijoin). After a query plan is chosen and executed, the final results are sent to the user, with the applied solution modifiers. 

\emph{SEnTriS} follows a similar logic, for query evaluation, with a few modifications to enhance security. The system works with data that is encrypted in a manner that allows for search and set operations to remain executable. Additionally, before each step of the pipeline, access checks are performed against the users' ephemeral access tokens, to ensure that only the right actors are performing actions within the system.

\emph{SSE} techniques are used during query evaluation: we generate trapdoors to search encrypted subjects, predicates and objects present in \emph{triple patterns}. With these, we build \emph{bindings tables}, which are used for the evaluation of complex \emph{graph patterns}

\gls{SEnTriS} is implemented using three different approaches:

\begin{itemize}
    \item \emph{\gls{SEnTriS} V0}: Operates over plaintext data and does not provide data confidentiality, but enforces role-based access control.

    \item \emph{\gls{SEnTriS} V1}: Data encryption retains elements from \emph{CryptDB} \autocite{Popa2011-zb, Popa2012-bw, Popa2011-ci}. Subjects, predicates and objects are encrypted in two layers. The outer layer is probabilistically encrypted, whereas the second is deterministically encrypted. When equality checks are required, we peel the outer layer and compare the values.

    \item \emph{\gls{SEnTriS} V2}: In this last implementation, we aim to reduce the leakage produced during evaluation of queries, by using an additive homomorphic encryption scheme. The value of a subject, predicate or object is mapped to an \emph{equality tags}. These values are homomorphically encrypted and support re-encryption. Trapdoors index the \emph{equality tags} and equality tags reference the probabilistically encrypted subjects, predicates and objects. The \emph{bindings tables} are built with \emph{re-encrypted equality tags}. When doing equality checks, we decrypt the difference between \emph{equality tags}.
\end{itemize}

In the remainder of this section we will: 1) begin by defining the core data elements and cryptographic tools; 2) specify the \gls{SPARQL} query evaluation protocol; 3) define the triple pattern \gls{SSE} schemes, and the adaptations done to the \gls{SPARQL} query evaluation protocol in order for it to work with encrypted data; 3) explain how we adapt the protocols to the \gls{M/M} setting; 5) define the various implementations of \gls{SEnTriS}; 6) define the security properties of \gls{SEnTriS}; 7) analyze the security and performance trade-offs of the various implementations.

\subsection{Core elements}

\paragraph{Triplestores} A \gls{RDF} dataset within \gls{SEnTriS} is represented as a single named triplestore. A triplestore $T = (triplestoreID,\; nodes, \; acp)$ is a key value store, where $triplestoreID$ is a unique string that identifies the dataset, $nodes = \{(idx_{0}, node_{0}), ... ,(idx_{n},\allowbreak node_{n})\}$ is a map of predicate, subject or object \emph{nodes}, $node \in I \cup B \cup L$, and each node is uniquely identified by its pattern index, $idx$, which is a hash of the \emph{triple pattern}. The $acp$ is the triplestore's \gls{ACP}.

This data structure is a simplified version of the one defined in section \ref{ch2:sec:semantic-web:sparql}, where we only represent the \emph{default} graph.

\paragraph{Triplestore Access Control Policy} A triplestore's \gls{ACP} $acp = (r,\; w,\; acrl)$, is a tuple where $r$ and $w$ are both collections of usernames and $acrl$ is a collection of \emph{access change requests}. An access change request $accessRequest = (requestID, \; username, \; write)$ is a tuple where $requestID$ is a \gls{UUID} (version 4) \autocite{Leach2005-rf}, $username$ identifies the user that this request affects and $write$ is a boolean, that specifies if the user desires write access (true) or read access (false).

\paragraph{Users} A user $u = (username, \; password, \; owned, \; role)$ is a tuple, where $username$ and $password$ are unique strings, $owned = [triplestoreID_0,...,triplestoreID_n]$ is a collection of user's owned triplestores identifiers and $role \in \{\textrm{BASIC}, \; \textrm{PRIVILEGED},\; \textrm{ADMIN}\}$.


\paragraph{Role Change Request} A role change request $roleRequest = (requestID, username, role)$ is a tuple where $requestID$ is a \gls{UUID} (version 4), $username$ identifies the user that this request affects and $role \in \{\textrm{BASIC}, \; \textrm{PRIVILEGED}, \textrm{ADMIN}\}$ specifies the desired new role.


\paragraph{Sessions} A user's session $s = (sessionID,\; username,\; ttl)$ is a tuple, where  $sessionID$ is a \gls{UUID} (version 4), $username$ belongs to a valid registered $user$ and $ttl \in N$ specifies the maximum time (in seconds) that a session can remain active, before expiring.


\paragraph{Locks} A lock is $l = (lockID,\; resourceID,\; ttl)$ is a tuple, where  $lockID$ is a \gls{UUID} (version 4), $resourceID$ is either a valid $username$ or $triplestoreID$ and $ttl \in N$ specifies the maximum time (in seconds) that a lock can remain valid, before expiring.


\paragraph{Ephemeral Access tokens} An ephemeral access token $at = (tokenID, \; ttl, \; username, \allowbreak sessionID, \; triplestoreID, \; lockID)$ is a tuple, where $tokenID$ is a \gls{UUID} (version 4), $ttl \in N$ specifies the remaining time (in seconds) that the token will stay active, $username$ is the identifier of the user who holds the token, $sessionID$ is the identifier of the token holder's active session, at the time of its creation, $triplestoreID$ is the identifier of the triplestore whose access is being requested and $lockID$ can either be $\bot$ (for \emph{read} access requests) or the identifier of the current lock, against the targeted triplestore (for \emph{write} access requests).


\paragraph{SPARQL queries} The definition of a \gls{SPARQL} query within \gls{SEnTriS} is equal to that provided in section \ref{ch2:sec:semantic-web:sparql}.


\paragraph{Query Execution Plans} A query execution plan is a tuple $plan=(jobs, \;execOrder)$, where $jobs = \{(jobID_0, \; job_0),...,(jobID_n, \; job_n)\}$ is a map that maps a query $job$ to their respective identifiers, which are \gls{UUID}s (version 4), and $execOrder$ is a sequence of job identifiers.


\paragraph{Query Jobs} A query job is an internal operation that is executed during the evaluation of a \gls{SPARQL} query. Query jobs can be of three types: \emph{basic}, \emph{unary} and \emph{binary} jobs. Unary and binary jobs represent operations that are executed against the \emph{bindings} of, respectively, one and two other jobs, whereas basic jobs are independent of other jobs' results. The execution of a job will always generate \emph{bindings}, even if it is an empty collection. These \emph{bindings} will be associated to the respective $jobID$.

There exist two basic jobs, $job_{SEARCH}$ and $job_{EMPTY}$.

The search job is a tuple $job_{SEARCH} = (jobID, \; varPattern, \; s, \; p, \; o)$, where $varPattern \in \{ \textrm{S}, \; \textrm{P}, \; \textrm{O}, \; \textrm{SP}, \; \textrm{SO}, \; \textrm{PO}, \; \textrm{SPO} \}$ and $s$, $p$, $o$ are $\bot$ or have the value of, respectively a subject, predicate or object of the \emph{triple pattern} being searched.

The empty result job is a tuple $job_{EMPTY} = (jobID, \; vars)$, where $vars = [var_0, ..., var_n]$ is a collection of variables.

There exist four unary jobs, which represent the operations that apply result modifiers. These jobs are tuples of type $job_{I} = ( jobID, \; jobID_{prev} )$, where $I \in \{ \textrm{PROJECT}, \; \textrm{ORDERBY}, \allowbreak \textrm{SLICE}, \; \textrm{DISTINCT} \}$ and dictates which solution modifier will be applied to the \emph{bindings} associated with $jobID_{prev}$.

Finally, there are four binary jobs, which represent set operations between \emph{binding}. These are tuples of type $job_{II} = ( jobID, \; jobID_l,  \; jobID_r )$, where $II \in \{ \textrm{JOIN}, \; \textrm{UNION}, \allowbreak \textrm{OPTIONAL},  \; \textrm{MINUS} \}$ and specifies which set operation will be executed between the \emph{bindings} of $jobID_l$ and $jobID_r$.


\paragraph{\emph{Bindings} tables} The output of a query job is a \emph{bindings} table. This data structure is a tuple $bindingsTable = (bindings, \; patterns)$ where $bindings = \{ (var_0, \; bindingsMap_0), ..., \allowbreak (var_n, \; bindingsMap_n)\}$ and $patterns = \{ (var_0, \; patternsMap_0), ...,  (var_n, \; patternsMap_n)\}$ are maps that map query variables, respectively, to \emph{bindings} and \emph{triple patterns} maps.

A \emph{bindings} map, $bindingsMap = \{ (binding_0, \; [pattternID_{0,0},...,patternID_{0,k}]),...,\linebreak(binding_n, \; [pattternID_{0,0},...,patternID_{n,k}])\}$, is a map where \emph{bindings} are mapped to a list of \emph{triple pattern} identifiers, which are \gls{UUID}s (version 4).

A \emph{triple patterns} map, $patternsMap = \{ (pattternID_0, \; binding_0),..., (patternID_n, \linebreak binding_n)\}$, is a map where \emph{triple pattern} identifiers are mapped to \emph{bindings}.


\paragraph{SPARQL Query Result} A \gls{SPARQL} query result is a tuple $sparqlResult = (bindings, \allowbreak distinct, \; ordered, \; sliced, \; sortConditions, \; length, \; \textrm{\textit{offset}})$ where $bindings = [pattern_0, ..., \allowbreak pattern_n]$ is a collection of \emph{triple patterns}, $ordered$, $distinct$ and $sliced$ are boolean values, $length$ and $\textrm{\textit{offset}}$ are integers and $sortConditions = [(var_0, \; dir_0), ..., (var_n, \; dir_n)]$ is a collection of variables and their respective sorting direction, which can either be $1$ or $-1$. A positive $dir$ represents an \emph{ascending} sorting whereas a negative $dir$ represents a \emph{descending} sorting condition.

Let $n \in \mathbb{Z}^{+}$ be the number of variables in a \gls{SPARQL} query, a triple pattern $pattern_i = \{(var_0, \; binding_{i,0}),...,(var_n, \; binding_{i,n})\}$ is a map where each variable is mapped to a $binding$. A $binding$ can be a plaintext value or $\bot$ for the particular case of $\textrm{OPTIONAL}$ queries.

\paragraph{SPARQL Final Query Result} A \gls{SPARQL} query result $sparqlResult$ is the result of evaluating a $WHERE$ clause. The final output of a query evaluation must respect the standards present in \cite{W3c2013-hd, W3c_undated-tw}. This final query result, $sparqlFinalResult$, will vary depending on the query type: 
\begin{itemize}
    \item \textit{SELECT}: it will be a query result in JSON format;
    \item \textit{ASK}: it will be a boolean;
    \item \textit{CONSTRUCT}: it will be a set of triples;
    \item \textit{DESCRIBE}: it will be a frequency table of the query variables;
    \item \textit{SPARQL UPDATE}: it will be either $\textrm{SUCCESS}$, or $\bot$ in case of failure;
\end{itemize}

\subsubsection{SEnTriS V1 and V2 specific}

We will denote $E(m)$ as the encryption of $m$, $O(m)$ as the obfuscation of $m$, mapping it to a random value.

\paragraph{Encrypted Triplestores} An encrypted triplestore $E(T) = (triplestoreID,\; E(nodes), \; acp)$ is a key value store, where $triplestoreID$ is a unique string that identifies the dataset, $E(nodes)$ is an encrypted map of predicate, subject or object \emph{nodes} and the $acp$ is the triplestore's \gls{ACP}. $E(nodes)$ will be defined in Section \ref{chapter3:definition:triple-pattern-encryption-schemes}


\paragraph{Secret Maps} A secret map $smap = (triplestoreID, \; secrets)$ is a tuple where $triplestoreID$ is the identifier of an existing triplestore, $secrets = \{(key_0, \; secret_0), ..., \allowbreak (key_n, \; secret_n)\}$ is a map and $key_i$ is a unique string that identifies $secret_i$, the serialized value of a secret parameter.

\paragraph{Secure Query Jobs} The only difference between jobs in \gls{SEnTriS} V1 and V2, to jobs in \gls{SEnTriS} V0, is that they operate over obfuscated variables and output encrypted \emph{bindings} tables.

Only the \emph{secure search job} is structurally different. It is a tuple $job_{SEC\_SEARCH} = (jobID, \; vars, \; searches)$, where $vars = [O(var_0), ..., \allowbreak O(var_n)]$ is a collection of obfuscated variables. During the query preparation phase, $searches = \{(O(var_0), \; keyword_0), ..., $\linebreak$ (O(var_n), \; keyword_n)\}$ is a map that maps variables to the keywords to be searched, whereas, in the query evaluation phase, $searches = \{(O(var_0), \; searchID_0), ...,  (O(var_n), \; searchID_n)\}$ is a map that maps variables to the identifiers of search results. These identifiers are \gls{UUID}s (version 4).

\paragraph{Randomized Query Execution Plans} A randomized query execution plan is a tuple $plan_{rnd}=(jobs, \;rndExecOrder)$, where $jobs = \{(jobID_0, \; job_0),...,(jobID_n, \; job_n)\}$ is a map that maps a query $job$ to their respective identifiers, which are \gls{UUID}s (version 4), and $rndExecOrder$ is a randomly chosen sequence of job identifiers, from the subset of valid execution sequences.

\paragraph{Randomized Prepared Query Execution Plans} After the query preparation phase, the $searches$ maps, in the secure search jobs, have been updated to map obfuscated variables to search identifiers. A plan where all search jobs have been updated like this is said to be \textit{prepared}.

\paragraph{Encrypted \emph{Bindings} tables} Similar to a \emph{bindings} table, an encrypted \emph{bindings} table is a tuple $E(BindingsTable) = (E(bindings), \; E(patterns))$ where $E(bindings) = \{ (O(var_0), \allowbreak E(bindingsMap_0)), ..., (O(var_n), \; E(bindingsMap_n))\}$ and $E(patterns) = \{ (O(var_0), \linebreak E(patternsMap_0)), ..., (O(var_n), \; E(patternsMap_n))\}$ are maps that map \emph{obfuscated} query variables, respectively, to encrypted \emph{bindings} and \emph{triple patterns} maps.

An encrypted \emph{bindings} map, $E(bindingsMap) = \{ (E(binding_0), \; [pattternID_{0,0},...,\linebreak patternID_{0,k}]),..., (E(binding_n), \; [pattternID_{0,0},...,patternID_{n,k}])\}$, is a map where encrypt-\linebreak\text{ed}  \emph{bindings} are mapped to a list of \emph{triple pattern} identifiers, which are \gls{UUID}s (version 4).

An encrypted \emph{triple patterns} map, $E(patternsMap) = \{ (pattternID_0, \; E(binding_0)),..., \allowbreak (patternID_n, \; E(binding_n))\}$, is a map where \emph{triple pattern} identifiers are mapped to encrypted \emph{bindings}.

We will define $E(binding)$ in Section \ref{chapter3:definition:triple-pattern-encryption-schemes} for both \gls{SEnTriS} V1 and \gls{SEnTriS} V2.


\paragraph{Encrypted SPARQL Query Result} An encrypted \gls{SPARQL} query result is a tuple \linebreak $E(sparqlResult) = (E(bindings), \allowbreak distinct, \; ordered, \; sliced, \; sortConditions, \; length, \; \textrm{\textit{offset}})$ where $E(bindings) = [E(pattern_0), ..., \allowbreak E(pattern_n)]$ is a collection of encrypted \emph{triple patterns}, $ordered$, $distinct$ and $sliced$ are boolean values, $length$ and $\textrm{\textit{offset}}$ are integers and $sortConditions = [(var_0, \allowbreak dir_0), ..., (var_n, \; dir_n)]$ is a collection of variables and their respective sorting direction, which can either be $1$ or $-1$. A positive $dir$ represents an \emph{ascending} sorting whereas a negative $dir$ represents a \emph{descending} sorting condition.

Let $n \in \mathbb{Z}^{+}$ be the number of variables (obfuscated) in a \gls{SPARQL} query, an encrypted triple pattern $E(pattern_i) = \{(O(var_0), \; E(binding_{i,0})),...,(O(var_n), \; E(binding_{i,n}))\}$ is a map where each obfuscated variable is mapped to an encrypted $binding$. An encrypted $binding$ can either have a value or be $\bot$ for the particular case of $\textrm{OPTIONAL}$ queries.

\subsection{Cryptographic Tools}

\subsubsection{Notation}

For the remainder of the document, the function $E(K, \; m, \; iv)$ denotes the encryption of message $m$, using the initialization vector $iv$ and key $K$. The notation $E(data)$ will represent uniquely that $data$ is encrypted. The function $D(K, \; ct)$ denotes the decryption of ciphertext $ct$, using key $K$. The keys used as input will identify the encryption scheme (e.g. $K_{A}$ implies the usage of the symmetric encryption scheme $A$ and $K_{B_{priv}}$ implies the usage of public key encryption scheme $B$).

\subsubsection{DGK Encryption}

\gls{SEnTriS} V2 can be implemented with any additive homomorphic encryption scheme, but we have chosen to use the \gls{DGK} encryption scheme \autocite{Damgard2007-km, Damgard2009-my, Franz2011-sm, Makkes2010-qb}. Compared to other schemes (e.g. Paillier \autocite{Paillier1999-js}) it produces smaller ciphertexts, and the reduced plaintext and ciphertext spaces result in large performance gains. It also permits us to employ more efficient equality checks, which is a very common operation when evaluating \gls{SPARQL} queries.

\begin{definition}[\textbf{DGK Encryption}]\label{ch3:definition:dgk} Let $k$, $t$, and $l$ be public parameters. Parameter $k$ determines the size of a \gls{RSA} modulus, $l$ is the bit length of the messages to encrypt and $t$ should be large enough so that exhaustive search is infeasible.

    \paragraph{Key Generation} Let $p$, $q$ be two primes, generated such that $n = pq$ is a $k$-bit \gls{RSA} modulus and that the $u$, $v_p$ and $v_q$ also exist and possess the following properties:

    \begin{changemargin}{1cm}{1cm}
        \begin{enumerate}
            \item $u$ is the smallest prime greater than $l + 2$
            \item $u$ is a divisor of both $p - 1$ and $q -1$
            \item $v_p$ is $t$-bit prime numbers that divides $p - 1$
            \item $v_q$ is $t$-bit prime numbers that divides $q - 1$
        \end{enumerate}
    \end{changemargin}


    Choose random values $g, h \in \mathbb{Z}^*_n$, so that:

    \begin{changemargin}{1cm}{1cm}
        \begin{enumerate}
            \item the multiplicative order of $h$ is $v_pv_q$
            \item $g$ is an element of order $uv_pv_q \in \mathbb{Z}^*_n$
        \end{enumerate}
    \end{changemargin}

    The message space is $M = \mathbb{Z}_u$ and the ciphertext space is $C = \{g^mh^r \; \textrm{mod} \; n \; | \; m \in \mathbb{Z}_u, \;  r \in \mathbb{Z}_{v_pv_q}\} \subseteq \mathbb{Z}^*_n$. The public and private keys are, respectively, $K_{DGK_{pub}} = (n,\;g,\;h,\;u)$ and $K_{DGK_{priv}} = (p,\;q,\;v_p,\;v_q)$.

    \paragraph{Probabilistic Encryption} An encryption $ct$ of $m$ is calculated as:
    \begin{equation}\label{chapter3:definition:dgk:equation:rnd-enc}
        ct = g^mh^r  \; \textrm{mod} \; n
    \end{equation}

    The value $r \in \mathbb{Z}_{v_pv_q}$ is only possible to calculate if we have access to private key. Knowing the public key, we instead determine the value of $r$ as a random $2.5t$-bit integer, so that $h^r$ is indistinguishable from a uniformly random element from the subgroup generated by $h$.

    \paragraph{Deterministic Encryption} To compute a deterministic encryption of $m$ we remove the randomization part, $h^r \; \textrm{mod} \; n$, from the calculations:
    \begin{equation}\label{chapter3:definition:dgk:equation:det-enc}
        ct = g^m \; \textrm{mod} \; n
    \end{equation}

    \paragraph{Re-encryption} To re-encrypt a probabilistic encryption of $m$ we sample a new random $2.5t$-bit integer $r_2$ and compute:
    \begin{align}\label{chapter3:definition:dgk:equation:re-enc}
        \begin{split}
            ct_2 &= ct \cdot h^{r_2} \; \textrm{mod} \; n \\
            &= g^mh^rh^{r_2} \; \textrm{mod} \; n \\
            &= g^mh^{r + r_2} \; \textrm{mod} \; n \\
        \end{split}
    \end{align}

    \paragraph{Decryption} The decryption of a ciphertext $ct$ is done by calculating:

    \begin{align}
        \begin{split}
            c^{v_pv_q} \; \textrm{mod} \; n &= (g^mh^r)^{v_pv_q} \; \textrm{mod} \; n \\
            &= (g^m)^{v_pv_q}(h^r)^{v_pv_q} \; \textrm{mod} \; n \\
            &= (g^m)^{v_pv_q}*1 \; \textrm{mod} \; n \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{because order of h is $v_pv_q$}\\
            &= (g^{v_pv_q})^m \; \textrm{mod} \; n
        \end{split}
    \end{align}

    Due to the value of $u$ being small, it is feasible to search for the calculated value in a look-up table with tuples of type $((g^{v_pv_q})^i \; \textrm{mod} \; n, \; i)$, for $0 < i < u$. Even without the look-up table, one can quickly determine if $ct$ is an encryption of 0, because $D(E(0)) = 1$:

    \begin{equation}\label{chapter3:definition:dgk:equation:zero-dec}
        c^{v_pv_q} \; \textrm{mod} \; n = (g^{v_pv_q})^m \; \textrm{mod} \; n \stackrel{\text{m = 0}}{=} (g^{v_pv_q})^0 \; \textrm{mod} \; n = 1 \; \textrm{mod} \; n = 1
    \end{equation}

    So far we have presented the decryption equations using all $p, q, v_p, v_q$ values, but we can utilize a faster decryption method. This method only requires the computation of modulo $p$ or $q$, because $m$ is uniquely determined from $c^{v_p} \; \textrm{mod} \; p$ or $c^{v_q} \; \textrm{mod} \; q$. For the remaining of the document we will only be considering this decryption method.

    \paragraph{Homomorphic property} The \gls{DGK} cryptosystem is homomorphic with respect to the addition operation ($+$). For $ct_1, ct_2 \in C$, there exists $m_1, m_2 \in M$ and $r_1, r_2 \in \mathbb{Z}_{v_pv_q}$, where $ct_1 = g^{m_1}h^{r_1}  \; \textrm{mod} \; n$ and $ct_2 = g^{m_2}h^{r_2}  \; \textrm{mod} \; n$ and Equation \ref{ch2:sec:homomorphic-encryption:def-eq-additive} holds:

    \begin{equation}
        \begin{split}
            E(m_1) * E(m_2) &= (g^{m_1}h^{r_1} \; \textrm{mod} \; n) * (g^{m_2}h^{r_2} \; \textrm{mod} \; n) \\
            &= ( g^{m_1}h^{r_1}  \; * g^{m_2}h^{r_2} ) \; \textrm{mod} \; n \\
            &= g^{(m_1 + m_2)}h^{(r_1 + r_2)} \; \textrm{mod} \; n \\
            &= E(m_1 + m_2)
        \end{split}
    \end{equation}

    \paragraph{Additional operations} Multiplication of an encrypted value by a positive constant $k$ is supported, just like in Equation \ref{ch2:sec:homomorphic-encryption:def-eq-additive-mul-positive}, by raising it to the power of $k$.

    Multiplication by a negative constant $c$ is done similarly to what is specified in Equation \ref{ch2:sec:homomorphic-encryption:def-eq-additive-mul-negative}, but instead of considering a plaintext space $\mathbb{Z}_n$, we consider $M = \mathbb{Z}_u$. As such, this operation requires raising the encrypted value to the power of $u - c$:

    \begin{equation}\label{chapter3:definition:dgk:equation:mul-negative}
        E(m)^{(u - c)} = E(-c*m)
    \end{equation}

    Given Equations \ref{ch2:sec:homomorphic-encryption:def-eq-additive-subtraciton} and \ref{chapter3:definition:dgk:equation:mul-negative}, subtraction of two ciphertexts is achieved by computing:

    \begin{equation}
        E(m_1 - m_2) = E(m_1 + (-1) * m_2) = E(m_1) * E(m_2)^{(u - 1)}
    \end{equation}

\end{definition}

To perform an equality check, between two plaintext messages, $m_1, m_2$, we compute the difference of their respective ciphertexts, $ct_1, ct_2$, and check if the result is 1 ($m_1 = m_2 \Leftrightarrow m_1 - m_2 = 0$):

\begin{equation}
    D(E(m_1) - E(m_2)) =  D(E(0)) = 1
\end{equation}

\subsection{Evaluation of SPARQL queries}

Now that we have defined all the core data structures and the cryptographic tools, we will proceed to explain how we evaluate \gls{SPARQL} queries. We will start by explaining the protocol for doing so over plaintext data, implemented in \gls{SEnTriS} V0. For simplicity, we will not take into consideration role based access control and permissions, as it will not alter in any way the protocol logic.

\subsubsection{Storage of triple patterns}

The first step when designing a solution that supports \gls{SPARQL}, is to define how we will store \emph{triple patterns}. The storage strategy must preserve semantics and ensure fast access during query evaluation.

To achieve this, when we index data, our goal is to preserve the relationships between subjects, predicates and objects (which might appear in queries' graph patterns). In the remainder of the document we will use \emph{node} as a common term to denominate \emph{subjects}, \emph{predicates} and \emph{objects}.

In \gls{SEnTriS} V0, we index a triple pattern, $t = (s, p, o)$ based on all of its possible variations within a \gls{SPARQL} query. For example, to answer a query where we need to search for $(?v, p, o)$ (the subject is a variable) we need an index $p,o \leftarrow [s_0, ..., s_n]$. With this index, all subjects are referenced by the predicate and object pairs in the triplestore. Applying this strategy to other possible patterns, we can index all triple pattern variations with the following 6 indexes:

\begin{align}
    \begin{split}
        &(p,o) \leftarrow [s_0, ..., s_n] \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(?v, p, o)$} \\
        &(s,o) \leftarrow [p_0, ..., p_n] \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(s, ?v, o)$} \\
        &(s,p) \leftarrow [o_0, ..., o_n] \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(?v, p, o)$} \\
        &s \leftarrow [(p,o)_0, ..., (p,o)_n] \;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(s, ?v_1, ?v_2)$} \\
        &p \leftarrow [(s,o)_0, ..., (s,o)_n] \,\;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(?v_1, p, ?v_2)$} \\
        &o \leftarrow [(p,s)_0, ..., (p,s)_n] \,\;\;\;\;\;\;\;\;\;\;\;\; \text{answers $(?v_1, ?v_2, o)$}
    \end{split}
\end{align}

We do not index patterns with 3 variables, as a join of two complementary indexes (e.g. $s$ with $(p, o)$) can answer this kind of queries. Nonetheless, because the encryption schemes of \gls{SEnTriS} V1 and V2 do not support this kind of triple pattern, we will not support this type of queries in V0 either. 

For patterns with 2 variables, the index consists in the concatenation of the pattern type (e.g, "PO"\space for indexes referencing predicate and object pairs) with the known value (e.g. a subject). This index will reference a set of node pairs, whose values we concatenate together.

For patterns with a single variable, the index is the concatenation of the pattern type (e.g. "S" \space for indexes referencing subjects) with the concatenation of the other two known values (e.g. predicate and object). This index references a set of individual nodes (e.g. subjects).

In Listing \ref{storage-engine-v0}, we define the logic for \gls{SEnTriS} V0 storage engine. We will assume that this engine executes write operations \emph{atomically}: it either executes all operations, or none.

\lstinputlisting[caption={Storage Engine for SEnTriS V0}, label={storage-engine-v0}]{Code/chapter3/storage-engine-v0.tex}

Notice that a search operation always produces a \emph{bindings table}. This data structure was designed to support the operations that need to be executed between intermediate collections of \emph{bindings}, during query evaluation. In Listing \ref{bindings-table-v0}, we show the underlying logic of a \emph{bindings table}.

\lstinputlisting[caption={Bindings Table in SEnTriS V0}, label={bindings-table-v0}]{Code/chapter3/bindings-table-v0.tex}

\subsubsection{SPARQL Query Evaluation Protocol}

Having defined how data is stored, we will now explain the protocol for evaluating \gls{SPARQL} queries.

\begin{definition}[\textbf{SPARQL Query Evaluation Protocol}]\label{ch3:definition:sparql-protocol}
    Let SPARQL-EVAL be a \emph{SPARQL} query evaluation protocol. It is a tuple $\textrm{SPARQL-EVAL} = (PlanQuery, \; EvaluateQuery, \allowbreak GenerateFinalResult)$ of 3 algorithms where:
    \begin{itemize}
        \item $\{(plan, \; construct, \; upload, \; delete), \; \bot\} \leftarrow \mathbf{PlanQuery(query)}$: Takes as input a \gls{SPARQL} query. If the query is well-formed, and all required operations are supported, outputs a query execution plan and three collections of triple patterns used while constructing the final query results: the $construct$, $upload$ and $delete$ collections. In case of any error, it outputs $\bot$.

        \item $\{sparqlResult, \bot\} \leftarrow \mathbf{EvaluateQuery(plan, \; triplestoreID)}$: Takes as input a query execution plan and the identifier of the triplestore to be queried. In case of success, outputs the query results, $\bot$ otherwise.
        
        \item $\{sparqlFinalResult, \bot\} \leftarrow \mathbf{GenerateFinalResult(triplestoreID, \; queryType, }$\linebreak$\mathbf{ variables, \; bindings, \; construct, \; upload, \; delete)}$: Takes as input a triplestore identifier, the query type, a collection of variables, the $bindings$ obtained from the evaluation of a $\textrm{WHERE}$ clause, and three collections of triple patterns. The $construct$ template will be used for generating the final result of a $\textrm{CONSTRUCT}$ query, whereas the $upload$ and $delete$ templates will be used to compute the triples to, respectively, upload and delete in the specified triplestore. In case of success outputs a $sparqlFinalResult$, $\bot$ otherwise.
    \end{itemize}
    SPARQL-EVAL is a distributed, two-party protocol: the query plan generation and final result generation are executed by a \emph{client} component; whereas the query evaluation and data updates are executed by a \emph{triplestore} component.
\end{definition}

\paragraph{Query Plan Generation} In \gls{SEnTriS} we created two abstractions to separate the logic for parsing a query from the logic for generating a query execution plan. Respectively, we denominate them as the \emph{SPARQL} \emph{query engine} and \emph{query planner}. The query engine will parse the query into processable operations. These operations will be consumed by the query planner, which will create a query execution plan.

The parsing of a \gls{SPARQL} query is out of the scope of this thesis. We've based our solution in \emph{Apache Jena} \autocite{The_Apache_Software_Foundation_undated-wk}. We will utilize the solution provided by the API to parse a query and decompose it into a collection of \emph{operations} (their equivalent abstraction to our \emph{query jobs}). We will not formally define these data structures, but we will assume that their implementation is correct and sufficient to generate valid query jobs. All of this logic is scoped to the query engine, and it is specified in Listing \ref{sparql-engine-pseudocode}.

\lstinputlisting[caption={\gls{SEnTriS} SPARQL Query Engine}, label={sparql-engine-pseudocode}]{Code/chapter3/sparql-engine.tex}

One can argue that the collection of \emph{Apache Jena} operations, created after parsing and compiling a query, are equivalent to a query execution plan: executing them by their precedence order, will generate a valid query evaluation. In Chapter \ref{cha:implementation}, we will explain why these are not compatible with our solution, and why we have decided to implement our custom query planner. The functioning of a simplified \gls{SEnTriS} V0 query planner (without any reasoning capabilities) is shown in Listing \ref{sparql-query-planner-pseudocode}.

\lstinputlisting[caption={\gls{SEnTriS} V0 SPARQL Query Planner (without reasoning capabilities)}, label={sparql-query-planner-pseudocode}]{Code/chapter3/sparql-planner-v0.tex}

Notice that the query planner generates a query execution plan where equi-joins are randomly ordered. The basis for this decision is to not incur in any bias, when evaluating performance, by always generating a deterministic execution order. Like the parsing of a \emph{SPARQL} query, optimization of its execution plan is out of the scope of this thesis, hence this approach.

We generate a random ordering of the equi-joins, by randomly sampling one of the valid execution orders. \gls{BGP}s are the only elements from a query that generate equi-joins. When we have visited all \gls{BGP} operations, we create an auxiliary graph where compatible equi-joins are neighbors. Two joins are \emph{compatible} if they have, at least, 1 common variable. We generate the execution order of the equi-joins, by exploring the graph and computing a path that contains each $job_{JOIN}$ once. To generate the random ordering, we randomly sample the origin of the path. After generating the execution order we add the query jobs to the execution plan.

\paragraph{Query Evaluation} Following the dispatcher pattern, we have created two other abstractions in \gls{SEnTriS}: the \emph{query execution}, and the \emph{query worker}. A query execution dispatches work, and a query worker processes pieces of work.

In the specific case of our query evaluation protocol, the query execution will process the query execution plan and dispatch the next available query job, to the worker. The worker has access to the stored data and will execute the assigned query job. The currently \gls{SEnTriS} only supports sequential dispatching of jobs to a single worker.

The dispatching of jobs continuous until there are no more jobs pending execution. After the last job is executed, the query result is generated and stays available to be retrieved as output. 

Listings \ref{query-execution-pseudocode} and \ref{query-worker-pseudocode} specify, respectively, the functioning of the query execution and query worker for \gls{SEnTriS} V0.

\lstinputlisting[caption={SEnTriS V0 Query Execution}, label={query-execution-pseudocode}]{Code/chapter3/query-execution.tex}

\lstinputlisting[caption={SEnTriS V0 Query Worker}, label={query-worker-pseudocode}]{Code/chapter3/query-worker.tex}

\paragraph{Generating Final Result \& Applying modifications} After the evaluation of a query's $\textrm{WHERE}$ clause, or the generation of the $delete$ and $update$ templates, there is still one final phase to execute. During this phase we generate the correct final query answer, which matches the query type, and update the triplestore data, upon the presence of an update query. Listings \ref{sparql-eval-client-pseudocode} and \ref{sparql-eval-triplestore-pseudocode} present, respectively, the logic of the SPARQL-EVAL \emph{client} and \emph{triplestore} components. These two components will serve as the base reference for the fully functional \gls{SEnTriS} components.

\lstinputlisting[caption={SPARQL-EVAL client}, label={sparql-eval-client-pseudocode}]{Code/chapter3/sparql-eval-client.tex}

\lstinputlisting[caption={SPARQL-EVAL triplestore}, label={sparql-eval-triplestore-pseudocode}]{Code/chapter3/sparql-eval-triplestore.tex}

\subsubsection{SPARQL Query Evaluation Protocol with Reasoning}

We will now explain how we modify the SPARQL-EVAL protocol, by applying \emph{query rewriting} techniques, during the generation of query execution plans, to retain reasoning capabilities. 

\begin{definition}[\textbf{SPARQL Query Evaluation Protocol with Reasoning}]\label{ch3:definition:sparql-protocol-reasoning}
    Let \linebreak SPARQL-EVAL-II be a \emph{SPARQL} query evaluation protocol that supports reasoning, via the application of query rewriting techniques.
    
    It is a tuple $\textrm{SPARQL-EVAL-II} = (PlanQuery, \; \textrm{SPARQL-EVAL}.EvaluateQuery, \linebreak \textrm{SPARQL-EVAL}.GenerateFinalResult)$ of 3 algorithms where:
    \begin{itemize}
        \item $\{(plan, \; construct, \; upload, \; delete), \; \bot\} \leftarrow \mathbf{PlanQuery(query, \; triplestoreID, }$\linebreak$\mathbf{ inference, \; expansionDepth, \; transitivityDepth)}$: Takes as input a \gls{SPARQL} query, a triplestore identifier, a boolean which indicates if the query needs to be rewritten and two integer parameters that control the rewriting depth. The $expansionDepth$ specifies the number of rounds and the $transitivityDepth$ specifies the number of transitivity relationships to considered. If the query is well-formed, and all required operations are supported, outputs a query execution plan and three collections of triple patterns used while constructing the final query results: the $construct$, $upload$ and $delete$ collections. In case of any error, it outputs $\bot$.
    \end{itemize}
    SPARQL-EVAL-II is a distributed, two-party protocol: the query plan generation and final result generation are executed by a \emph{client} component; whereas the query evaluation and data updates are executed by a \emph{triplestore} component.
\end{definition}

\paragraph{Storage of Schema triples} We will refer to the subset of triples containing nodes that belong to the Tbox as \emph{schema triples}. Storage of such triples will not require indexation by the pattern type. Furthermore, in \gls{SEnTriS} we will only support a single set of schema triples per triplestore, which can either be uploaded or deleted, but not modified. In Listings \ref{storage-engine-schema-v0} and \ref{sparql-eval-ii-triplestore-pseudocode} we specify, respectively, the modifications to \gls{SEnTriS} V0 storage engine and the extra operations of the SPARQL-EVAL-II triplestore component, required to support the storage and retrieval of such triples.

\lstinputlisting[caption={Storage of schema triples}, label={storage-engine-schema-v0}]{Code/chapter3/storage-engine-schema-v0.tex}

\lstinputlisting[caption={SPARQL-EVAL-II triplestore (extra operations)}, label={sparql-eval-ii-triplestore-pseudocode}]{Code/chapter3/sparql-eval-ii-triplestore.tex}

\paragraph{Query Rewriting} To rewrite the queries we first fetch the schema triples, and then instantiate the triplestore's ontology. In \gls{SEnTriS}, an ontology is an abstraction where we map class and property axioms to nodes. 

Using the \emph{Apache Jena} API \autocite{The_Apache_Software_Foundation_undated-wk} we create an ontological model. After the generation of the model, we load it with the schema triples. Once the triples are loaded, we iterate over all the classes and map each one to equivalent classes, subclasses, restrictions and intersections they are part of.

In the next step, we iterate over all properties and map each one to their subproperties, equivalent and inverse properties. We also map the properties to two boolean parameters which are true if, respectively, the property is symmetric and transitive.

All of this information is stored in adequate auxiliary data structures which will be accessible by the query planner. Listing \ref{ontology-pseudocode} provides the specification of a triplestore's ontology in \gls{SEnTriS} V0.

\lstinputlisting[caption={SEnTriS V0 Ontology}, label={ontology-pseudocode}]{Code/chapter3/ontology.tex}

During the creation of the query execution plan, when visiting a \gls{BGP}, the query planner consults the ontology and searches for information related with the nodes present in the triple patterns. This information is used to expand the \gls{BGP} with new patterns. The logic to rewrite queries is scoped to the query planner. All the modifications required to the query planner and client are depicted in Listings \ref{query-rewriting-v0} and \ref{sparql-eval-ii-client-pseudocode} respectively. 

\lstinputlisting[caption={Query Plan Generation with Query Rewriting}, label={query-rewriting-v0}]{Code/chapter3/query-rewriting.tex}

\lstinputlisting[caption={SPARQL-EVAL-II client (modified operations)}, label={sparql-eval-ii-client-pseudocode}]{Code/chapter3/sparql-eval-ii-client.tex}

Optimization and development of a fully functional query rewriting technique is out of the scope of this thesis. The system can not evaluate property paths, functions and nested queries, which limits our available solutions. We focused on trying to support a few expansions related with: subclasses, equivalent classes, restrictions, intersections, subproperties, equivalent and inverse properties, property transitivity and symmetry. Furthermore, we assume that the data which will be queried is consistent with the ontology provided. 

We expand the triple patterns in a leveled approach. We decided to implement this solution with parametrized expansion and transitivity depth because it offers control over the expansion of the query, which impacts performance: the expansion tends to introduce additional operations.

To conclude, this approach may also introduce duplicates into the results of a query, so we are required to use the DISTINCT modifier. 

\subsubsection{Supported Queries}

The two protocols defined for query evaluation do not offer full \gls{SPARQL} support. All versions of \gls{SEnTriS} will only be capable of evaluating a subset of queries with a few restrictions. Queries will solely be evaluated against the default graph. Only the DISTINCT, ORDERBY and LIMIT modifiers are allowed. Finally, graph patterns must not contain property paths, nested subqueries queries, graph selection, nor triple patterns composed solely by variables. Following the grammar used in Definition \ref{ch2:definition:sparql-graph-pattern}, a graph pattern will be defined as:

\begin{equation}
    \begin{split} 
    P ::= \;\; & tp \;\;\vert\;\; P_1\; .\; P_2 \;\;\vert\;\; P_1\; \textrm{UNION}\; P_2 \;\;\vert\;\; P_1\; \textrm{OPTIONAL}\; P_2 \;\;\vert\;\; P_1\; \textrm{MINUS}\; P_2
    \end{split}
\end{equation}

All query types will be supported apart from the LOAD and CLEAR, which will be implemented with dedicated operations to create, upload and delete a triplestore.


\subsection{Encryption of triple patterns}
\label{chapter3:definition:triple-pattern-encryption-schemes}

In \gls{SEnTriS} V0 query evaluation is executed with plaintext data, but in the other two versions the protocol will be adapted to work with encrypted data. As such, the encryption schemes must encrypt data in such a way that it retains its functionality. For the subset of supported queries, data must be searchable via triple patterns and allow the equality operation, essential for the execution of equi-joins.

\subsubsection{Triple Pattern SSE scheme}

\begin{definition}[\textbf{Triple Pattern SSE Scheme}]\label{ch3:definition:sse-sentri-det}
    Let $\textrm{SSE-TP}$ be triple pattern \gls{SSE} scheme. It is defined as a tuple $\textrm{SSE-TP} = (CreateTriplestore, \; Upload, \; Delete, \; UpdateTriplestore, \allowbreak DeleteTriplestore, \; SearchTriplestore, \; SearchSchema, \; Setup, \; ReInitialize, \; Encrypt, \linebreak Decrypt, \; Trapdoor, \; Search, \; PrepareUpload,  PrepareDelete, \; Update, \; DeleteAll)$ of 17 algorithms where:
    \begin{itemize}
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{CreateTriplestore(triplestoreID, \; [t_0, t_1, ..., t_n])}$: Takes as input a triplestore identifier and a set of \emph{ABox} triples. Outputs SUCCESS if the triplestore is successfully created and encrypted, $\bot$ otherwise.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{Upload(triplestoreID, \; [t_0, t_1, ..., t_n], \; schema)}$: Takes as input a triplestore identifier, a set of triples and a boolean parameter, which is set to true if the triples are schema triples. If triples are successfully encrypted and uploaded outputs SUCCESS, $\bot$ otherwise.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{Delete(triplestoreID, \; [t_0, t_1, ..., t_n])}$: Takes as input a triplestore identifier, a set of triples and a boolean parameter, which is set to true if the triples are schema triples. If triples are successfully deleted from the triplestore outputs SUCCESS, $\bot$ otherwise.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{UpdateTriplestore(triplestoreID, \; triples_{delete}, \; triples_{upload})}$: Takes as input a triplestore identifier, a set of triples to upload and a set of triples to delete. Outputs SUCCESS if the triples to delete are successfully deleted and the triples to upload are successfully encrypted and uploaded, $\bot$ otherwise.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{DeleteTriplestore(triplestoreID, \; schema)}$: Takes as input the triplestore identifier and a boolean parameter, which is set to true if only the schema triples should be deleted. Outputs SUCCESS if the requested data is deleted, $\bot$ otherwise.
        \item $\{[node_0, ..., node_n], \; \bot\} \leftarrow \mathbf{SearchTriplestore(triplestoreID, \; triplePatterns)}$: Takes as input the triplestore identifier and a collection of triples patterns. Searches the triplestore for matches against the triple patterns.Outputs a collection of nodes, or $\bot$ if an error occurred.
        \item $\{[t_0, t_1, ..., t_n], \; \bot\} \leftarrow \mathbf{SearchSchema(triplestoreID)}$: Takes as input the triplestore identifier. Searches the encrypted triplestore for its schema triples. Outputs a collection of triples, or $\bot$ if an error occurred.
        \item $\lambda \leftarrow \mathbf{Setup()}$: Takes no input. Outputs a set of newly generated security parameters.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{ReInitialize(\lambda)}$: Takes as input a set of security parameters. In case of success, sets the security parameters, updates the state of the scheme and outputs SUCCESS, $\bot$ otherwise.
        \item $E(nodes) \leftarrow \mathbf{Encrypt([t_0,..., t_n], \; schema)}$: Takes as input a set of triples and a boolean parameter, which is set to true if the triples are schema triples. Parses the data received and outputs a map of encrypted nodes, indexed by trapdoors.
        \item $node \leftarrow \mathbf{Decrypt(E(node), \; schema)}$: Takes as input an encrypted node and a boolean parameter, which is set to true if the encrypted node belongs to the schema triples. Decrypts the node and outputs the plaintext value.
        \item $trapdoor_{n,i} \leftarrow \mathbf{Trapdoor(keyword_n, \; i)}$: Takes as input a keyword, and an integer. Generates and outputs the trapdoor that corresponds to the $i^{th}$ trapdoor of the keyword.
        \item $[E(node_0), ..., E(node_n)] \leftarrow \mathbf{Search(triplestoreID, \; [trapdoor_0, ..., trapdoor_n])}$: Takes as input the triplestore identifier and a collection of trapdoors. Searches the encrypted triplestore for matches against the received trapdoors. Outputs a collection of encrypted nodes, where there was a match.
        \item $\{uploadID, \bot\} \leftarrow \mathbf{PrepareUpload(triplestoreID, \; E(nodes))}$: Takes as input the triplestore identifier and a map of encrypted nodes, indexed by trapdoors. Temporarily saves the operation. Outputs the identifier of the uploads to be applied upon the execution of an SSE-TP.$Update$ operation, $\bot$ in case of an error.
        \item $\{deletionID, \bot\} \leftarrow \mathbf{PrepareDelete(triplestoreID, \; [trapdoor_0, ..., trapdoor_n])}$: Takes as input the triplestore identifier and a collection of trapdoors, whose entries should be deleted. Temporarily saves the operation. Outputs the identifier of the deletions to be applied upon the execution of an SSE-TP.$Update$ operation, $\bot$ in case of an error.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{Update(triplestoreID, \; [uploadID_0, ..., uploadID_n], \; [deletionID_0,}$ \linebreak $\mathbf{ ..., deletionID_n])}$: Takes as input the triplestore identifier, a collection of upload identifiers, and a collection of deletion identifiers. Outputs SUCCESS if the uploads and deletions are successfully executed, $\bot$ otherwise.
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{DeleteAll(triplestoreID)}$: Takes as input the triplestore identifier. Outputs SUCCESS if the triplestore data is deleted, $\bot$ otherwise.
    \end{itemize}
    SSE-TP is a distributed, two-party protocol: search and storage of encrypted data is scoped to the \emph{triplestore} component; whereas triplestore creation, and data encryption and decryption is scoped to the \emph{client} component.
\end{definition}

\subsubsection{Trapdoor generation}

Following the \gls{SSE} approach, in \gls{SEnTriS} V1 and V2 we do not reference nodes using plaintext indexes. Instead, we generate \emph{trapdoors} by deterministically encrypting the value of the indexes that would reference the node. To enforce the property that every trapdoor must uniquely reference a single node, we deterministically encrypt the index value using an incremental \emph{initialization vector}. This initialization vector is equal to the observed frequency of the index keyword.

Trapdoors generated with $iv = 0$ always reference the latest value of the index keyword frequency.

For each index, the respective trapdoors are encrypted using a unique key. This key is build from a master key, for the context of the index keyword.


\begin{definition}[\textbf{Trapdoor Generation}]\label{ch3:definition:trapdoor-generation}
    Let $keyword_n$ be the keyword that references node $n \in \textrm{ABox}$. Let $K_{MASTER}$ be a master key for a deterministic encryption scheme and $K_{keyword_n}$ be the derived key for the context of $keyword_n$. Finally, let $i$ be the observed frequency for $keyword_n$, and $trapdoor_{n,i}$ be the trapdoor that is built from $keyword_n$, after $i - 1$ observations of the term:
    \begin{equation}
        trapdoor_{n, i} = E(K_{keyword_n}, \; keyword_n, \; i)
    \end{equation}
\end{definition}

In the case of the trapdoors that reference schema nodes, we will also base the solution in the previous indexing scheme, but instead of using a fixed schema keyword, we will randomly generate it for each triplestore.

\begin{definition}[\textbf{Schema Node Trapdoor Generation}]\label{ch3:definition:schema-trapdoor-generation}
    Let $keyword_{schema}$ be a randomly generated keyword, specific to a triplestore $t$. Let $n \in \textrm{TBox}$ be a schema node, $K_{MASTER}$ be a master key for a deterministic encryption scheme and $K_{schema}$ be the derived key for the context of $keyword_{schema}$. Finally, let $i$ be the observed frequency for $keyword_{schema}$, and $trapdoor_{schema,i}$ be the trapdoor that is built from $keyword_{schema}$, after $i - 1$ observations of the term:
    \begin{equation}
        trapdoor_{schema,i} = E(K_{schema}, \; keyword_{schema}, \; i)
    \end{equation}
\end{definition}

We also reference with trapdoors the frequencies of a triple pattern, which are necessary for computing deletions.

\begin{definition}[\textbf{Triple Frequencies Trapdoor Generation}]\label{ch3:definition:schema-triple-freq-generation}
    Let $keyword_{s}$, $keyword_{p}$ and $keyword_{o}$ be the keywords that reference the nodes belonging to a triple pattern $t = (s, p, o)$. Let $keyword_{t} = \textrm{SPO} \; \vert\vert \; keyword_{s} \; \vert\vert \; keyword_{p} \; \vert\vert \; keyword_{o}$ be the keyword that references $t$, $K_{MASTER}$ be a master key for a deterministic encryption scheme and $K_{keyword_{t}}$ be the derived key for the context of $keyword_{t}$. Finally, let $i$ be the observed frequency for $keyword_{t}$, and $trapdoor_{t,i}$ be the trapdoor that is built from $keyword_{t}$, after $i - 1$ observations of the term:
    \begin{equation}
        trapdoor_{t,i} = E(K_{keyword_{t}}, \; keyword_{t}, \; i)
    \end{equation}
\end{definition}

Finally, uniquely to \gls{SEnTriS} V2, trapdoors generated with $iv = 0$ for the context of the parsed value of a node, reference its \emph{equality tag}. We will define equality tags in Section \ref{ch3:encryption-definition}

\subsubsection{Initial setup \& re-initialization of scheme} During initialization, we generate all the encryption keys and security parameters of the scheme. These vary between versions the two versions. Re-initialization is required because the scheme is stateful, the state exclusive to each triplestore, and we might need to execute multiple operations against different triplestores.

\paragraph{SEnTriS V1} The security parameters correspond to $\lambda = (K_{MASTER}, \; K_{RND}, \; K_{DET}, \; iv_{DET}, \allowbreak keyword_{schema})$ where, $K_{MASTER}$, $K_{RND}$ and $K_{DET}$ are secret keys, $iv_{DET}$ is a randomly generated initialization vector and $keyword_{schema}$ is a freshly generated a \gls{UUID} (version 4).

\paragraph{SEnTriS V2} The security parameters correspond to $\lambda = (K_{MASTER}, \; K_{RND}, \; K_{DGK_{priv}}, \linebreak K_{DGK_{pub}}, \; iv_{DET}, \; keyword_{schema})$ where, $K_{MASTER}$ and $K_{RND}$ are secret keys, $K_{DGK_{priv}}$ and $K_{DGK_{pub}}$ are, respectively, a \gls{DGK} private and public key, $iv_{DET}$ is a randomly generated initialization vector and $keyword_{schema}$ is a freshly generated a \gls{UUID} (version 4).

\subsubsection{Triple Pattern encryption and decryption} \label{ch3:encryption-definition} The implementation of the scheme in both versions is very similar: the generation of trapdoors does not vary; the encryption of schema nodes and frequencies belonging to the triple patterns are probabilistically encrypted. They only differ on the encryption of nodes belonging to the \emph{ABox}. This common logic is specified in Listing \ref{encryption-scheme-base}.

\lstinputlisting[caption={Base logic for the encryption scheme}, label={encryption-scheme-base}]{Code/chapter3/scheme-base.tex}

\paragraph*{SEnTriS V1} We follow the \emph{CryptDB} \autocite{Popa2011-zb, Popa2012-bw, Popa2011-ci} approach and encrypt \emph{ABox} nodes in two layered onions $E(node) = E(K_{RND}, E(K_{DET}, \; node, \; iv_{DET}))$. The outer layer is probabilistically encrypted, whereas the second is deterministically encrypted. When equality checks are required, we peel the outer layer and compare the values. The version specific functions for \gls{SEnTriS} V1 are shown in Listing \ref{encryption-scheme-deterministic}.

\lstinputlisting[caption={Triple Pattern Encryption for SEnTriS V1}, label={encryption-scheme-deterministic}]{Code/chapter3/scheme-deterministic.tex}

\paragraph*{SEnTriS V2} \emph{ABox} nodes are mapped to \emph{equality tags}. We generate equality tags by using an injective function $eqTag: (I \cup L \cup V) \rightarrow \mathbb{Z}^{+}$ which maps the value of a node to a unique positive integer.

Trapdoors index encrypted equality tags and encrypted equality tags index encrypted \emph{ABox} nodes. Respectively, nodes are probabilistically encrypted $E(node) = E(K_{RND}, \; node )$ and equality tags are homomorphically encrypted using the \gls{DGK} cryptosystem \autocite{Damgard2007-km, Damgard2009-my, Franz2011-sm, Makkes2010-qb} $E(eq_{tag}) = E(K_{DGK_{pub}}, \; eqTag(node))$. When doing equality checks, we decrypt the equality tags and compare the values. The version specific functions for \gls{SEnTriS} V2 are shown in Listing \ref{encryption-scheme-dgk}.

\lstinputlisting[caption={Triple Pattern Encryption for SEnTriS V2}, label={encryption-scheme-dgk}]{Code/chapter3/scheme-dgk.tex}

\subsubsection{Storage, searches \& updates} When we create a triplestore, we initialize its encryption protocol by generating the security parameters. We then parse its \emph{ABox} triple patterns, encrypt and upload them. The schema triples can be uploaded in separate upload operations. The triplestore creation and encryption operations are handled by the client. 

Due to the high volumes of data that can be processed during write operations, these algorithms were designed to be executed in batches. Consequently, upon receiving the encrypted data pertaining to a triplestore, the triplestore component will only store it in temporary entries. These are fully persisted when a SSE-TP.$Update$ operation is executed. We will continue to assume that the storage engine operations are atomic. In Listings \ref{sse-tp-storage-engine} and \ref{sse-tp-triplestore-component} we specify, respectively, the storage engine and the triplestore component logic.

\lstinputlisting[caption={SSE-TP Storage Engine}, label={sse-tp-storage-engine}]{Code/chapter3/sse-tp-storage-engine.tex}

\lstinputlisting[caption={SSE-TP Triplestore Component}, label={sse-tp-triplestore-component}]{Code/chapter3/sse-tp-triplestore.tex}

The search operation is a two-step process in \gls{SEnTriS} V1. Firstly, we parse the triple patterns and generate their keyword frequency trapdoors (where $i = 0$) and search for these values. Upon receiving the encrypted values, we decrypt the frequencies, generate trapdoors accordingly and execute the final search operation.

In \gls{SEnTriS} V2, the results of the second search yield a collection of encrypted equality tags. We need an extra communication round to retrieve the encrypted nodes, referenced by the encrypted equality tags. This extra communication round is only required for search over the \emph{ABox}, so the algorithm is exactly the same in both version.

Listings \ref{sse-tp-client-search-v1} and \ref{sse-tp-client-search-v2} show the search logic in the client component for versions V1 and V2, respectively.

\lstinputlisting[caption={Searches in SSE-TP Client (SEnTriS V1)}, label={sse-tp-client-search-v1}]{Code/chapter3/sse-tp-client-search-v1.tex}

\lstinputlisting[caption={Searches in SSE-TP Client (SEnTriS V2)}, label={sse-tp-client-search-v2}]{Code/chapter3/sse-tp-client-search-v2.tex}

In \gls{SEnTriS} V1, to execute deletions, we fetch the frequencies of the triple patterns, generate their trapdoors and send a deletion request for data referenced by those trapdoors. In the case of \gls{SEnTriS} V2, we are required to fetch the equality tags, because data referenced by them needs to be deleted as well. The deletion requests are sent to the triplestore, where they are temporarily saved.

To upload new data, firstly, we fetch the frequencies of the triple patterns keywords. Additionally, in \gls{SEnTriS} V2, we will need to fetch the equality tags. Upon receiving all the contextual data, we decrypt it, and set the encryption scheme to the right context. We then issue deletion requests for any duplicated data, encrypt the new triple patterns and send the upload requests to the triplestore component.  

Update operations are the only way to actually persist changes in the triplestores, so they are always executed at the end of an upload or deletion request. To ensure that triplestores always stay in a consistent state after subsequent modifications, deletions are processed first, and then the updated values are stored.

Listings \ref{sse-tp-client-writes-v1} and \ref{sse-tp-client-writes-v2} show the underlying logic of write operations in the client component. For version V2, we only highlight the algorithms that differ between versions.

\lstinputlisting[caption={Write operations in SSE-TP Client (SEnTriS V1)}, label={sse-tp-client-writes-v1}]{Code/chapter3/sse-tp-client-writes-v1.tex}

\lstinputlisting[caption={Write operations in SSE-TP Client (SEnTriS V2)}, label={sse-tp-client-writes-v2}]{Code/chapter3/sse-tp-client-writes-v2.tex}

\subsection{Evaluation of SPARQL queries over encrypted data}

We will now explain how we join the SPARQL-EVAL-II with the SSE-TP protocols, and create a protocol capable of evaluating \gls{SPARQL} queries that works with encrypted triplestores.

\begin{definition}[\textbf{Secure SPARQL Query Evaluation Protocol}]\label{ch3:definition:secure-sparql-protocol}
    Let SSE-SPARQL-EVAL be a \emph{SPARQL} query evaluation protocol, that works with SSE-TP encrypted triplestores. It is a tuple $\textrm{SSE-SPARQL-EVAL} = (\textrm{SSE-TP}.ReInitialize, \; \textrm{SSE-TP}.Encrypt, \; \textrm{SSE-TP}.Decrypt, \linebreak \textrm{SSE-TP}.PrepareUpload, \; \textrm{SSE-TP}.PrepareDelete, \; \textrm{SSE-TP}.Update, \; \textrm{SSE-TP}.Trapdoor, \linebreak \textrm{SSE-TP}.Search, \; PlanQuery, \; PrepareQuery, \; PrepareSearch, \; PrepareBindings, \linebreak EvaluateQuery, \; GenerateFinalResult)$ of 8 algorithms where:
    \begin{itemize}        
        \item $\{(plan_{rnd}, \; construct, \; upload, \; delete), \; \bot\} \leftarrow \mathbf{PlanQuery(query, \; triplestoreID, }$\linebreak$\mathbf{ inference, \; expansionDepth, \; transitivityDepth)}$: Takes as input a \gls{SPARQL} query, a triplestore identifier, a boolean which indicates if the query needs to be rewritten and two integer parameters that control the rewriting depth. The $expansionDepth$ specifies the number of rounds and the $transitivityDepth$ specifies the number of transitivity relationships to considered. If the query is well-formed, and all required operations are supported, outputs a randomized query execution plan and three collections of triple patterns used while constructing the final query results: the $construct$, $upload$ and $delete$ collections. In case of any error, it outputs $\bot$.
         
        \item $\{ preparedPlan_{rnd}, \; \bot \} \leftarrow \mathbf{PrepareQuery(triplestoreID, \; plan_{rnd})}$: Takes as input a \linebreak triplestore identifier and a randomized query execution plan. For each distinct $job_{SEC\_SEARCH}$ in the plan a SSE-SPARQL-EVAL.$PrepareSearch$ will be triggered. Upon receiving the $searchID$, the respective $job_{SEC\_SEARCH}$ keywords will be updated accordingly. Outputs a randomized prepared query execution plan. In case of any error, outputs $\bot$.
        
        \item $\{ searchID, \; \bot \} \leftarrow \mathbf{PrepareSearch(triplestoreID, \; mask, \; [trapdoor_0, ..., trapdoor_n])}$: \linebreak Takes as input a triplestore identifier, a masking value and a collection of trapdoors. Upon finding the values, these will be re-encrypted, if $mask \neq \bot$. Finally, a SSE-SPARQL-EVAL.$PrepareBindings$ operation is executed. Outputs the value received from the SSE-SPARQL-EVAL.$PrepareBindings$ operation.
        
        \item $\{ searchID, \; \bot \} \leftarrow \mathbf{PrepareBindings([E(binding_0),...,E(binding_n)])}$: Takes as input a collection of encrypted bindings. Generates a fresh \gls{UUID} (version 4) and persists the received values in memory, associating them with the generated identifier. Outputs the identifier, or $\bot$ if an error occurs.

        \item $\{E(sparqlResult), \; \bot\} \leftarrow \mathbf{EvaluateQuery(preparedPlan_{rnd}, \; K)}$: Takes as input a randomized prepared query execution plan and a secret key. Executes the query plan and outputs the encrypted \gls{SPARQL} results. If there is an error, outputs $\bot$. 
        
        \item $\{sparqlFinalResult, \bot\} \leftarrow \mathbf{GenerateFinalResult(triplestoreID, \; queryType, }$\linebreak$\mathbf{ variables, \; bindings, \; construct, \; upload, \; delete)}$: Takes as input a triplestore identifier, the query type, a variables map, the $bindings$ obtained from the evaluation of a $\textrm{WHERE}$ clause, and three collections of triple patterns. The $construct$ template will be used for generating the final result of a $\textrm{CONSTRUCT}$ query, whereas the $upload$ and $delete$ templates will be used to compute the triples to, respectively, upload and delete in the specified triplestore. In case of success outputs a $sparqlFinalResult$, $\bot$ otherwise.
    \end{itemize}
    SSE-SPARQL-EVAL is a distributed, three-party protocol: search and storage of encrypted data is scoped to the \emph{triplestore} component; encryption and decryption of data, query plan generation, preparation and final result generation is scoped to the \emph{client} component; the evaluation of the query execution plans are scoped to a secure computation \emph{proxy} component.
\end{definition}

\subsubsection{Query Preparation Phase} In this phase we generate the query execution plan. We adapt the solution  the SPARQL-EVAL-II protocol in two ways. Firstly, we obfuscate the variables. Secondly, to introduce extra randomness in the query plan structure, the ontology and the expansion algorithm are randomized. Listing \ref{ontology-rnd} and \ref{secure-sparql-planner} specify the modifications to the \gls{SPARQL} planner and ontology logic.

\lstinputlisting[caption={SSE-SPARQL-EVAL ontology}, label={ontology-rnd}]{Code/chapter3/ontology-rnd.tex}

\lstinputlisting[caption={SSE-SPARQL-EVAL SPARQL Planner}, label={secure-sparql-planner}]{Code/chapter3/sparql-planner-secure.tex}


Upon the generation of a randomized query execution plan, the client component processes the secure search jobs and requests the necessary SSL-SPARQL-EVAL.$PrepareSearch$ operations. In \gls{SEnTriS} V2, a masking value is also computed and sent alongside each of the previous requests. 

After receiving the collections of trapdoors, the triplestore component matches them against the triplestore's data. The bindings found are sent to the computation proxy, to be temporarily persisted and used during the query evaluation phase. In \gls{SEnTriS} V2, before sending the data to the proxy, the bindings are re-encrypted, using the $mask$ value.

After receiving the search identifiers, the client substitutes the search jobs keywords with the respective values. After all search jobs have been successfully updated, the plan is said to be prepared and ready to be sent to the proxy component. Listings \ref{sse-sparql-client-v1} and \ref{sse-sparql-client-v2} show the logic of the client for \gls{SEnTriS} V1 and V2 respectively.  

\lstinputlisting[caption={Query Preparation in SSE-SPARQL-EVAL Client (SEnTriS V1)}, label={sse-sparql-client-v1}]{Code/chapter3/sse-sparql-client-v1.tex}

\lstinputlisting[caption={Query Preparation in SSE-SPARQL-EVAL Client (SEnTriS V2)}, label={sse-sparql-client-v2}]{Code/chapter3/sse-sparql-client-v2.tex}

Listings \ref{sse-sparql-triplestore} and \ref{sse-sparql-triplestore-storage} show the logic of the triplestore component and how search values are blinded in the storage engine.

\lstinputlisting[caption={SSE-SPARQL-EVAL Triplestore}, label={sse-sparql-triplestore}]{Code/chapter3/sse-sparql-triplestore.tex}

\lstinputlisting[caption={SSE-SPARQL-EVAL Triplestore Storage Engine}, label={sse-sparql-triplestore-storage}]{Code/chapter3/sse-sparql-triplestore-storage-engine.tex}

Finally, listings \ref{sse-sparql-proxy-prepare-bindings} and \ref{sse-proxy-save-bindings} provide the logic for the proxy component and how the bindings are temporarily stored. We have only highlighted the algorithms that differ from the previous protocols.

\lstinputlisting[caption={Query Preparation in SSE-SPARQL-EVAL Proxy}, label={sse-sparql-proxy-prepare-bindings}]{Code/chapter3/sse-sparql-proxy-prepare-bindings.tex}

\lstinputlisting[caption={Persisting Bindings in SSE-SPARQL-EVAL Proxy Storage Engine}, label={sse-proxy-save-bindings}]{Code/chapter3/sse-sparql-proxy-save-bindings.tex}

\subsubsection{Query Evaluation Phase} Query evaluation is done in the computation proxy and retains most of the logic present in the two SPARQL-EVAL protocols. The only difference is in the query worker. This worker will have access to a secret key and will operate with encrypted bindings tables. Listing \ref{sse-sparql-proxy-query-eval} shows the query evaluation logic in the proxy component. Listing \ref{query-worker-v1} specifies the modifications to the query worker. 

\lstinputlisting[caption={Query Evaluation in SSE-SPARQL-EVAL Proxy}, label={sse-sparql-proxy-query-eval}]{Code/chapter3/sse-sparql-proxy-evaluate-query.tex}

\lstinputlisting[caption={SSE-SPARQL-EVAL Query Worker}, label={query-worker-v1}]{Code/chapter3/query-worker-v1.tex}

As seen in Listing \ref{construct-bindings-tables-sentris-v1}, in \gls{SEnTriS} V1, before constructing the bindings tables, the probabilistic encryption layer of the bindings will be decrypted. The deterministic values allow the bindings tables to operate as normal. 

\lstinputlisting[caption={Constructing Bindings Tables (SEnTriS V1)}, label={construct-bindings-tables-sentris-v1}]{Code/chapter3/sse-sparql-proxy-construct-bindings-tables-v1.tex}

In \gls{SEnTriS} V2, the values remain probabilistic. Consequently, equality checks, and set operations that require this operation, will have to be keyed as illustrated in Listing \ref{query-worker-v2}.

\lstinputlisting[caption={Keyed set operations in SSE-SPARQL-EVAL Query Worker (SEnTriS V2)}, label={query-worker-v2}]{Code/chapter3/query-worker-v2.tex}

Before constructing the encryption tables we will to group the values, keeping track of pattern indexes but retaining one unique value, per distinct group of equal equality tags. This will enable us to construct the encryption tables without altering their data structures. The non-deterministic nature of the data will result in an increase of complexity for internal searches, as they will have to be done using sequential scans. In Chapter \ref{cha:implementation}, we will explain how we mitigate this impact using parallelism.

Listings \ref{construct-bindings-tables-sentris-v2} and \ref{bindings-tables-sentris-v2} provide the logic for the construction of bindings tables, in \gls{SEnTriS} V2, and their respective modifications.

\lstinputlisting[caption={Constructing Bindings Tables (SEnTriS V2)}, label={construct-bindings-tables-sentris-v2}]{Code/chapter3/sse-sparql-proxy-construct-bindings-tables-v2.tex}

\lstinputlisting[caption={Bindings Tables (SEnTriS V2)}, label={bindings-tables-sentris-v2}]{Code/chapter3/bindings-table-v2.tex}

\subsubsection{Decryption of SPARQL query results} Upon receiving the results from the query evaluation, the client component will decrypt them and construct the final results. This final step will require an extra communication round for \gls{SEnTriS} V2, because the returned values are encrypted equality tags. These will need to be unmasked, and used to search for the encrypted nodes. 

After the decryption of the results, any modifications required are executed. Listings \ref{sse-sparql-final-results-v1} and \ref{sse-sparql-final-results-v2} illustrate this logic for \gls{SEnTriS} V1 and V2, respectively.

\lstinputlisting[caption={Generating Final Results in SSE-SPARQL-EVAL (SEnTriS V1)}, label={sse-sparql-final-results-v1}]{Code/chapter3/sse-sparql-client-v1-final-results.tex}

\lstinputlisting[caption={Generating Final Results in SSE-SPARQL-EVAL (SEnTriS V2)}, label={sse-sparql-final-results-v2}]{Code/chapter3/sse-sparql-client-v2-final-results.tex}

\subsection{The M/M setting}
\label{ch3:sec:solution-definition:mm-setting}

So far we have specified the building blocks for the secure evaluation of \gls{SPARQL} queries, but \gls{SEnTriS} is more than a query evaluation protocol. It is also a triplestore management and sharing protocol: users can create, manage and share encrypted and non-encrypted triplestores with varying degrees of access. 
To adapt the previous protocols to the \gls{M/M} setting we will require an identity and access management protocol, with secret sharing between authorized users for \gls{SEnTriS} V1 and V2.

Authentication will be validated with \emph{sessions}, management of access to triplestores will be achieved with the definition of triplestore \glsxtrlong{ACL}s and strict \emph{permission rules}. Finally, we will rely on a lock-based access strategy, and \emph{ephemeral access tokens}, to ensure that concurrent access, when executing unsafe operations (writes), is done in a controlled and safe manner.

\begin{definition}[\textbf{Identity and Access Management Protocol with Secret Sharing}]\label{ch3:definition:iam-protocol}
    Let IAM be an identity and access management protocol between authenticated users, where temporary access to triplestores is granted using ephemeral access tokens. It is a tuple of $\textrm{IAM} = ()$ of 45 algorithms where:
    \begin{itemize}
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{init()}$: initializes the protocol by registering a default admin user. Outputs $\textrm{SUCCESS}$ if the user is created and registered, $\bot$ otherwise.

        \item $\{cookie, \bot\} \leftarrow \mathbf{createSession(username, \; password)}$: takes as input a pair of user credentials. In case of valid credentials, creates a new session and outputs a $cookie$, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{save(username, \; password)}$: takes as input a pair of user credentials. Outputs $\textrm{SUCCESS}$ if a new user is registered, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{delete(cookie, \; username)}$ takes as input a cookie and the username of the user to delete. Outputs $\textrm{SUCCESS}$ if the user is successfully deleted, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow  \mathbf{issueGrantRoleRequest(cookie, \; username, \;issuer, \; role)}$: \linebreak takes as input a cookie, the usernames of the target and issuer of the role changing request and the desired role. Outputs $\textrm{SUCCESS}$ if the request is successfully persisted, $\bot$ otherwise.

        \item $[roleRequest_0,...,roleRequest_n] \leftarrow  \mathbf{getPendingRoleRequests(cookie, \; user{\text -}}$\linebreak $\mathbf{name)}$: takes as input a cookie, and the request issuer username. Outputs the collection of pending role changing requests.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{processRoleRequest(cookie, \; username, \; target, \; requestID,}$ \linebreak $\mathbf{decision)}$: takes as input a cookie, the username of the processing request's issuer, the username of the role change target, the request identifier and the issuer's decision. Outputs $\textrm{SUCCESS}$ if the role changing request is successfully processed, according to the provided decision, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{createTriplestoreAccessPolicy(cookie, \; issuer, \; triplesto{\text -}}$\linebreak $\mathbf{reID)}$: takes as input a cookie, the username of the request's issuer and a triplestore identifier. Outputs $\textrm{SUCCESS}$ if the triplestore access policy is sucessfully generated, $\bot$ otherwise.

        \item $[triplestoreID_0, ..., triplestoreID_n]  \leftarrow \mathbf{getTriplestores(cookie, \; issuer, \; owns, }$\linebreak$\mathbf{read, \; write)}$: takes as input a cookie, the username of the request's issuer and three boolean filter parameters. Outputs a collection of triplestore identifiers: if $owns$ is \emph{true}, returns identifiers of \emph{owned} triplestores; if $read$ is \emph{true}, returns identifiers of triplestores that the user has \emph{read} access; if $write$ is \emph{true}, returns identifiers of triplestores that the user has \emph{write} access;

        \item $\{msg, \bot\} \leftarrow \mathbf{getUsersWithAccess(cookie, triplestoreID, \; write, \; at)}$: takes as input a cookie, a triplestore identifier, a boolean filter parameter and an access token. If the access token has sufficient permissions, outputs a message with the specified triplestore's \emph{owner} username and the collection of \emph{read} and \emph{write} access (if the $write$ parameter is \emph{true}) users' usernames, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{changeTriplestoreOwner(cookie, \; triplestoreID, \; target, \; at)}$: takes as input a cookie, a triplestore identifier, a username and an access token. Outputs $\textrm{SUCCESS}$ if the ownership of the specified triplestore is successfully transfered to the $target$ user, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteTriplestoreAccessPolicy(cookie, \; triplestoreID, \; at)}$: \linebreak takes as input a cookie a triplestore identifier and an access token. Outputs $\textrm{SUCCESS}$ if the specified triplestore access policy is successfully deleted, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{grantAccess(cookie, \; triplestoreID, \; target,\; write, \; at)}$: takes as input a cookie, a triplestore identifier, a username, a boolean parameter and an access token. Outputs $\textrm{SUCCESS}$ if the $target$ user is successfully granted \emph{read}, or \emph{write} (if $write$ parameter is \emph{true}), access against the specified triplestore, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{revokeAccess(cookie, \; triplestoreID, \; target,\; write, \; at)}$: \linebreak takes as input a cookie, a triplestore identifier, a username, a boolean parameter and an access token. Outputs $\textrm{SUCCESS}$ if the $target$ user \emph{read}, or \emph{write} (if $write$ parameter is \emph{true}), access against the specified triplestore is successfully revoked, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{issueAccessRequest(cookie, \; triplestoreID, \; target,\; write)}$: \linebreak takes as input a cookie, a triplestore identifier, a username, a boolean parameter and an access token. Outputs $\textrm{SUCCESS}$ if an access change request, for the $target$ user, granting \emph{read} or \emph{write} (if the $write$ parameter is \emph{true}) permissions against the specified triplestore is sucessfully issued, $\bot$ otherwise.

        \item $\{[accessRequest_0, ..., accessRequest_n], \bot\} \leftarrow \mathbf{getPendingAccessRequests(cookie, }$\linebreak$\mathbf{ triplestoreID, \; at)}$: takes as input a cookie, a triplestore indentifier and an access token. Outputs the collection of the triplestore's pending access change requests, if the specified access token has sufficient permissions, $\bot$ otherwise.


        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{processAccessRequest(cookie, \; triplestoreID, \; requestID,}$\linebreak$\mathbf{decision, \; at)}$: takes as input a cookie, a triplestore identifier, an access change request identifier, a boolean parameter and an access token. Outputs $\textrm{SUCCESS}$ if the specified request is successfully processed according to the provided $decision$ parameter, $\bot$ otherwise.

        \item $\{tokenID, \bot\} \leftarrow \mathbf{createAccessToken(cookie, \; triplestoreID, \; target)}$: takes as input a cookie, a triplestore identifier, and a username. Outputs the newly generated token identifier, if the $target$ user has sufficient permissions against the specified triplestore, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteAccessToken(cookie, \; triplestoreID, \; at)}$: takes as input a cookie, a triplestore identifier, and an access token. Outputs $\textrm{SUCCESS}$ if the token is successfully deleted, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{checkReadAccess(triplestoreID, \; at)}$: takes as input a triplestore identifier and an access token. Outputs $\textrm{SUCCESS}$ if the token grants \emph{read} access to the specified triplestore, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{checkWriteAccess(triplestoreID, \; at)}$: takes as input a triplestore identifier and an access token. Outputs $\textrm{SUCCESS}$ if the token grants \emph{write} access to the specified triplestore, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{checkOwnerAccess(triplestoreID, \; at)}$: takes as input a triplestore identifier and an access token. Outputs $\textrm{SUCCESS}$ if the token grants \emph{owner} access to the specified triplestore, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{checkIfActive(at)}$: takes as input a triplestore identifier and an access token. Outputs $\textrm{SUCCESS}$ if the token is active (the associated session or the token has not expired), $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{acquireTriplestoreLock(cookie, \; at)}$: takes as input a cookie and an access token. Outputs $\textrm{SUCCESS}$ if the token is successfully associated with a newly generated lock against the triplestore affected by the token, $\bot$ otherwise.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{releaseTriplestoreLock(cookie, \; at)}$: takes as input a cookie and an access token. Outputs $\textrm{SUCCESS}$ if the lock associated with the token is successfully unlocked, $\bot$ otherwise.

        \item $\{cookie, \bot\} \leftarrow \mathbf{auth(username, \; password)}$: takes as input a pair of user credentials, executes $\textrm{IAM}.auth(username, \; password)$ and outputs the result of the algorithm.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{registerUser(username, \; password)}$: takes as input a pair of user credentials, executes $\textrm{IAM}.registerUser(username, \; password)$ and outputs the result of the algorithm.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteUser(cookie, \linebreak username)}$: takes as input a cookie and the username of the user to delete, executes $\textrm{IAM}.deleteUser(cookie, \; username)$ and outputs the result of the algorithm.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow  \mathbf{issueUpgradeRequest(cookie, \; username)}$: takes as input a cookie, and a username. Executes $\textrm{IAM}.issueGrantRoleRequest(cookie, \linebreak username, \; username, \; \textrm{PRIVILEGED})$ and outputs the result of the algorithm.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow  \mathbf{issueDowngradeRequest(cookie, \; username)}$: takes as input a cookie, and a username. Executes $\textrm{IAM}.issueGrantRoleRequest(cookie, \linebreak username, \; username, \; \textrm{BASIC})$ and outputs the result of the algorithm.

        \item $[roleRequest_0,...,roleRequest_n] \leftarrow  \mathbf{listPendingRoleRequests(cookie, \; user{\text -}}$\linebreak $\mathbf{name)}$: takes as input a cookie, and the request issuer username. Executes $\textrm{IAM}.getPendingRoleRequests(cookie, \; username)$ and outputs the result of the algorithm.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{processRoleRequest(cookie, \; username, \; target, \; requestID,}$ \linebreak $\mathbf{decision)}$: takes as input a cookie, the username of the processing request's issuer, the username of the role change target, the request identifier and the issuer's decision. Executes $\textrm{IAM}.processRoleRequest(cookie, \; username, \; target, \allowbreak requestID, \;  decision)$ and outputs the result of the algorithm.

        \item $\{[triplestoreID_0, ..., triplestoreID_n], \bot\} \leftarrow \mathbf{listTriplestores(cookie, \; issuer, }$\linebreak$\mathbf{ write, \; read, \; owns)}$: takes as input a cookie, the username of the request's issuer and three boolean filter parameters. Executes $\textrm{IAM}.listTriplestores(cookie, \allowbreak issuer, \; write, \; read, \; owns)$ and outputs the result of the algorithm, or $\bot$ in case of error.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{updateTriplestoreOwner(cookie, \;  triplestoreID, \; issuer, }$\linebreak $\mathbf{ target)}$: takes as input a cookie, a triplestore identifier, a username and an access token. Creates an access token by executing ... . In case of success, tries to lock the triplestore with ... . If it fails to lock the resource, deletes the access token by executing ... . If the lock is successfully acquired, executes ... and retrieves the response of the algorithm. Finally, releases the lock and deletes the access token, with ... followed by ... , and outputs the retrieved response.

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{grantAccess(cookie, \;  triplestoreID, \; issuer, \; target, \; write)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{revokeAccess(cookie, \;  triplestoreID, \; issuer, \; target, \; write)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{issueAccessRequest(cookie, \;  triplestoreID, \; issuer, \; write)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{listPendingAccessRequests(cookie, \;  triplestoreID, \; issuer)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{processAccessRequest(cookie, \;  triplestoreID, \; issuer, }$\linebreak $\mathbf{ requestID, \; decision)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{listUsersWithAccess(cookie, \;  triplestoreID, \; issuer, \; write)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{createTriplestore(cookie, \; triplestoreID)}$:

        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteTriplestore(cookie, \; issuer, \;  triplestoreID, \; inference)}$:
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{createSecrets(triplestoreID, \; secrets, \; at)}$:
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteSecrets(triplestoreID, \; at)}$:
        \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{getSecrets(triplestoreID, \; at)}$:
    \end{itemize}
    IAM is a distributed, multi-party protocol between a \emph{client}, a \emph{IAM provider} and a \emph{vault} component. The client is the public component, that users interact with. The IAM provider validates sessions, manages user data and triplestore access policies, stores role and permission change requests, validates permission rules, emits session cookies and ephemeral access tokens, validates access requests and controls the locking and unlocking of resources. The vault component is responsible for storage of secrets.
\end{definition}

\subsubsection{Session Validation} 

Authentication is the process where we verify the identity of a user. We do so, by checking that the provided credentials, during login, match the ones saved within our system. To avoid having to verify this condition in every request sent by the user, we use \emph{sessions}. The session is generated upon a successful login request. A valid session will prove the identity of the user sending the requests. In \gls{SEnTriS}, an authenticated user needs to be valid (has been registered within the system) and hold an active session.

\begin{definition}[\textbf{Authentication Check}]\label{ch3:definition:authentication} Let $IsAuth(user)$ be a predicate that evaluates to $true$ if a $user$ is authenticated, $\bot$ otherwise. Given session $s_i = (sessionID,\; username_s,\; ttl)$ and the valid $user_u = (username_u, \; password_u,\; role_u,\; owned_u)$, the authentication check will be upheld by verifying that the following statement is true: 
    \begin{equation}
        IsAuth(user_u) \Leftrightarrow \exists_{s} (username_u = username_s \; \wedge \; ttl > 0)
    \end{equation}
\end{definition}

The login request is initiated by the client component, which will redirect the provided credentials to the IAM provider, responsible for the emission of a cookie, holding the session data. Listings ... and ... provide the authentication logic for the client and IAM provider components respectively.

\subsubsection{Enforcing Access Permissions} 

Permissions in \gls{SEnTriS} are logical rules which dictate whether a user is allowed to execute a certain operation, or not. These rules are continuously checked during execution.

\begin{definition}[\textbf{Permission Rules}]\label{ch3:definition:permissions}
    Let $HasAccess(user, \; triplestore, \allowbreak type)$ be a predicate that evaluates to true if  $user$ has access of a certain $type$, to the specified $triplestore$, $\bot$ otherwise. Let $CanGrantAccess(user, \; type, \;  triplestore, \; target)$, and $CanRevokeAccess(user, \allowbreak type, \; triplestore, \; target)$ be predicates that evaluate to $true$, respectively, if a $user$ can grant or revoke access, of a certain $type$, to the specified $triplestore$ and specified $target$ user, $\bot$ otherwise.

    Let $CanCreateTriplestore(user, \; triplestore)$, $CanDeleteTriplestore(user, triplestore)$, $CanChangeOwnership(user, \; triplestore, \; target)$ be predicates that evaluate to $true$, respectively, if a $user$ can create, delete, or transfer ownership to a $target$ user, of the specified $triplestore$, $\bot$ otherwise.

    Finally, let $CanGrantRole(user, \; role, \; target)$, $CanRevokeRole(user, \; role, \; target)$ be predicates that evaluate to $true$, respectively, if a $user$ can grant, or revoke, the specified $target$ user's $role$, $\bot$ otherwise.

    Given $type \in \{\textrm{READ},\textrm{WRITE},\textrm{OWNER}\}$, triplestore $T_i = (triplestoreID_i,\; nodes_i, \; acp_i = (r_i, w_i))$, users $user_u = (username_u, \; password_u, \; role_u, \; owned_u)$ and $user_t = (username_t, \allowbreak role_t, \; password_t, \; owned_t)$, the validation of permission rules will consist in ascertaining that the following statements are true:

    \begin{align}
            Can&GrantAccess(user_u, \; type, \; T_i, \; user_t) \Leftrightarrow IsAuth(user_u) \; \wedge \; \exists_{user_t} \; user_u \neq user_t \\
            & \wedge \; (role_u = \textrm{ADMIN} \; \vee \; triplestoreID_i \in \; owned_u ) \notag\\
            & \notag \allowdisplaybreaks \\ 
            Can&RevokeAccess(user_u, \; type, \; T_i, \; user_t) \Leftrightarrow IsAuth(user_u) \\
            & \wedge \; \exists_{user_t} \; user_u \neq user_t \; \wedge \; (role_u = \textrm{ADMIN} \; \vee \; triplestoreID_i \in \; owned_u ) \notag \\
            & \notag \allowdisplaybreaks \\ 
            Has&Access(user_u, \; type, \; T_i) \Leftrightarrow role_u = \textrm{ADMIN} \; \vee \; triplestoreID_i \in \; owned_u  \\
            & \vee \; (type = \textrm{READ} \; \wedge \; username_u \in r_i) \; \vee \; (type = \{\textrm{READ}, \; \textrm{WRITE}\} \; \wedge \; username_u \in w_i) \notag \\
            & \vee \; (type = \textrm{OWNER} \; \wedge \; triplestoreID_i \in \; owned_u) \notag \\
            & \notag \allowdisplaybreaks \\
            Can&CreateTriplestore(user_u, \; T_i) \Leftrightarrow IsAuth(user_u) \\
            & \wedge \; role_u \in \{\textrm{ADMIN}, \; \textrm{PRIVILEGED}\} \; \wedge \; \nexists\;_{triplestore} \;\; triplestoreID = triplestoreID_i \notag \\
            & \notag \allowdisplaybreaks \\
            Can&DeleteTriplestore(user_u, \; T_i) \Leftrightarrow IsAuth(user_u) \\
            & \wedge \; role_u = \textrm{ADMIN} \; \wedge \; triplestoreID_i \in \; owned_u \notag \\
            & \notag \allowdisplaybreaks \\
            Can&ChangeOwnership(user_u, \; T_i, \; user_t) \Leftrightarrow IsAuth(user_u) \; \wedge \; \exists_{user_t} \; user_u \neq user_t \\
            & \wedge \; (role_u = \textrm{ADMIN} \; \vee \; triplestoreID_i \in \; owned_u ) \; \wedge \; role_t \neq \textrm{BASIC} \notag \\
            & \notag \allowdisplaybreaks \\
            Can&GrantRole(user_u, \; role, \; user_t) \Leftrightarrow IsAuth(user_u) \\
            & \wedge \; role \neq role_t \; \wedge \; ((\exists_{user_t} \; user_u \neq user_t \; \wedge \; role_u = \textrm{ADMIN}) \notag \\
            & \vee \; (user_u = user_t \; \wedge \; (role_u = \textrm{ADMIN} \; \vee \; role = \textrm{BASIC}))) \notag \\
            & \notag \allowdisplaybreaks \\
            Can&RevokeRole(user_u, \; role, \; user_t) \Leftrightarrow IsAuth(user_u) \\
            & \wedge \; role \neq role_t \; \wedge \; ((\exists_{user_t} \; user_u \neq user_t \; \wedge role_u = \textrm{ADMIN}) \; \vee \; user_u = user_t) \notag
    \end{align}
\end{definition}

The IAM provider is responsible for managing the roles of users and validating permission rules. The management of access permissions against triplestores is done with the usage of \gls{ACL}s. The logic for these validations is specified in Listing ... .

\subsubsection{Lock-based Concurrent Access}

Concurrent operations can target users or triplestores. The correctness of concurrent write and read operations is achieved using a \emph{lock-based} approach. Resources (users and triplestores) can be in a \emph{locked} or \emph{unlocked} state. Write operations can only be requested against \emph{unlocked} resources, which remain locked while the operation is being executed. Consequently, read operations can occur simultaneously, but write operations are sequential, against the same resource.

\emph{Locks} are enough for requests targeting users, because the execution context is always the same. Operations can only be executed by users, against themselves, or by \emph{admin} users, and the origin of the requests is always scoped to a \emph{Client} component.

Operations against triplestores are more complex. The query evaluation pipeline is distributed between different components. As such, whenever the execution context changes we must verify if the author of the request has sufficient permissions to execute the operations being requested.

To fulfill such requirements, every request contains an \emph{access token}. Permission checks are executed, by sending this token to a verifier component, in the case of \gls{SEnTriS} it will be the \emph{\gls{IAM} Provider}. The \emph{\gls{IAM} Provider} verifies the validity of the token, and if its associated permissions are correct.

Queries that contain write predicates, will require that their target remains \emph{locked}, during their evaluation. As such, an \emph{access token} can be associated with a \emph{lock}. The targetted triplestore will remain in a \emph{locked} state until its \emph{access token} is deleted or the \emph{lock} expires.

\begin{definition}[\textbf{Concurrent Lock-based Access Checks}]\label{ch3:definition:concurrent-access}
    Let $IsLocked(resource, \; l)$ and $CanLock(triplestore, \; at)$ be predicates that evaluate to $true$ if, respectively, $resource$ is in a $locked$ state associated with the lock $l$ and $triplestore$ can transition to a $locked$ state, and associate the generated lock with access token $at$, $\bot$ otherwise.

    Let $CanExecuteOperations(triplestore, \; at, \; type)$ be a predicate that evaluates to $true$ if the access token $at$ grants sufficient permissions, against the target $triplestore$, to allow operations, of the specified access $type$, to be executed in the current context of execution, $\bot$ otherwise.

    Given $type \in \{\textrm{READ},\textrm{WRITE},\textrm{OWNER}\}$, triplestore $T_i = (triplestoreID_i,\; nodes_i, \;  acp_i = (r_i, w_i))$, lock $l_a = (lockID_a,\; resourceID_a,\; ttl_a)$, lock $l_b = (lockID_b,\; resourceID_b,\; ttl_b)$, access token $at_k = (tokenID_k, \; username_k,\; ttl_k, \; sessionID_k, \; triplestoreID_k, \; lockID_k)$, user $user_u = (username_u, \; password_u, \; role_u, \; owned_u)$ and session $s_j = (session_{ID_j}, \; username_j, \allowbreak ttl_j)$, let the resources $resource_a$ and $resource_b$ be identified by, respectively, $resourceID_a$ and $resourceID_b$. The concurrent access checks will consist on verifying if the following statements are true:
    \begin{align}
        IsL&ocked(resource_a, \; l_a) \Rightarrow \exists_{l_a}\exists_{resource_a}(ttl_a \; > \; 0) \\
        & \wedge \; \nexists_{l_b}(resourceID_b \; = \; resourceID_a \; \wedge \; lockID_a \neq \; lockID_b \; \wedge \; ttl_b \; > \; 0) \notag \\
        & \wedge \; \nexists_{l_b}(resourceID_b \; \neq \; resourceID_a \; \wedge \; lockID_a = \; lockID_b \; \wedge \; ttl_b \; > \; 0) \notag \\
        & \notag \allowdisplaybreaks \\
        Ca&nLock(T_i, \; at_k) \Leftrightarrow \nexists_l(IsLocked(T_i, \; l)) \; \wedge \; triplestoreID_i \; = \; triplestoreID_k \\
        & \wedge \; username_k \; = \; username_u \; \wedge \; username_j \; = \; username_u \; \wedge \; ttl_j \; > \; 0  \notag \\
        & \wedge \; ttl_k \; > \; 0 \; \wedge \; HasAccess(user_u, \; type, \; T_i) \; \wedge \; type \in \{\textrm{WRITE}, \; \textrm{OWNER}\}  \notag \\
        & \notag \allowdisplaybreaks \\
        Ca&nExecuteOperations(T_i, \; at_k, \; type) \Leftrightarrow triplestoreID_i \; = \; triplestoreID_k \\
        & \wedge \; username_k \; = \; username_u \; \wedge \; username_j \; = \; username_u \; \wedge \; ttl_j \; > \; 0  \notag \\
        & \wedge \; ttl_k \; > \; 0 \; \wedge \; HasAccess(user_u, \; type, \; T_i) \; \wedge \; ( type \; = \; \textrm{READ}  \notag \\
        & \vee \; (type \in \{\textrm{WRITE}, \; \textrm{OWNER}\} \; \wedge \; IsLocked(T_i, \; l_a) \; \wedge \; lockID_a \; = \; lockID_k)) \notag
    \end{align}
\end{definition}

The emission of locks and ephemeral access tokens is also scoped to the IAM provider. We specify this logic in Listing ... .

\subsubsection{Issuing Role Change and Access Requests} 

\subsubsection{Triplestore Creation, Management and Sharing} 

\subsection{Definition of SEnTriS}
\label{ch3:sec:solution-definition:formal-definition}

\subsubsection{SEnTriS V0}

\subsubsection{SEnTriS V1 \& V2}

\begin{definition}[\textbf{SEnTriS V1}]\label{ch3:definition:sentri-v1}
    \gls{SEnTriS} V1 extends functionalities of \gls{SEnTriS} V0 to work with encrypted data. It is a collection of 5 protocols $(\textrm{SEnTriS-V0.IAM}, \; \textrm{CLIENT}, \; \textrm{VAULT}, \allowbreak \textrm{TRIPLESTORE}, \; \textrm{PROXY})$ where:

    \begin{itemize}
        \item CLIENT is a tuple of 20 polynomial time algorithms. Like in version V0, this protocol centralizes all functionalities of \gls{SEnTriS} and abstracts the communication logic with the other protocols. It retains the same functionalities, and only differs from version V0, in the algorithms for triplestore management and querying. We will only define these algorithms, and assume the others to be equal as their previous definition:

              \begin{itemize}
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{createTriplestore(cookie, \; triplestoreID)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{uploadTriplestoreContent(cookie, \; issuer, \;  triplestoreID, }$ \linebreak $\mathbf{ schema, \; syntax, \; content)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteTriplestore(cookie, \; issuer, \;  triplestoreID, \; inference)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{fetchTriplestoreSchema(cookie, \;  triplestoreID, \; inference)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{answerSPARQLQuery(cookie, \;  triplestoreID, \; issuer, }$\linebreak $\mathbf{ query, \; inference, \; transitivityDepth, \; expansionDepth)}$:

              \end{itemize}

        \item VAULT is a tuple of 3 polynomial time algorithms. These algorithms are responsible for ...:
              \begin{itemize}
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{create(triplestoreID, \; secrets, \; at)}$:
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{delete(triplestoreID, \; at)}$:
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{get(triplestoreID, \; at)}$:
              \end{itemize}

        \item TRIPLESTORE is a tuple of 6 polynomial time algorithms. These algorithms are responsible for ...:
              \begin{itemize}
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{upload(triplestoreID, \; encNodes, \; at)}$:
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{delete(triplestoreID, \; at)}$:
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{delete(triplestoreID, \; trapdoors, \; at)}$:
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{update(triplestoreID, \; deletions, \; uploads, \; at)}$:
                  \item $\{encBindings, \bot\} \leftarrow \mathbf{search(triplestoreID, \; trapdoors, \; at)}$:
                  \item $\{searchID, \bot\} \leftarrow \mathbf{prepareQuery(triplestoreID, \; trapdoors, \; at)}$:
              \end{itemize}

        \item PROXY is a tuple of 2 polynomial time algorithms. These algorithms are responsible for saving the outputs of TRIPLESTORE.$search(...)$ operations, preparing this data for query evaluation, executing \gls{SPARQL} query plans and outputting valid query results:
              \begin{itemize}
                  \item $\{searchID, \bot\} \leftarrow \mathbf{prepareQuery(encBindings, \; at)}$: takes as input a collection of encrypted bindings and an access token. If the token is valid, and grants sufficient permissions, saves the bindings in memory and associates the data with a newly generated $searchID$, a \gls{UUID} (version 4). In case of success, outputs the identifier, $\bot$ otherwise.

                  \item $\{sparqlResult, \bot\} \leftarrow \mathbf{answerSPARQLQuery(plan, K_{RND}, \; at)}$ takes as input a collection of a \gls{SPARQL} query plan, a secret key and an access token. If the token is valid, and grants sufficient permissions, executes the query plan and generates a query result. In case of success outputs the result, $\bot$ otherwise.
              \end{itemize}
    \end{itemize}
\end{definition}

\begin{definition}[\textbf{SEnTriS V2}]\label{ch3:definition:sentri-v2}
    \gls{SEnTriS} V2 is a collection of 5 protocols $(\textrm{TRIPLESTORE}, \allowbreak \textrm{SEnTriS-V0.IAM}, \textrm{CLIENT}, \; \textrm{SEnTriS-V1.VAULT}, \; \textrm{PROXY})$:

    \begin{itemize}
        \item CLIENT is a tuple of 20 polynomial time algorithms. Like in previous versions, the protocol centralizes all functionalities of \gls{SEnTriS} and abstracts the communication logic with the other protocols. It retains the same functionalities, and only differs from version V1, in the algorithms for triplestore management and querying. We will only define these algorithms, and assume the others to be equal as their previous definition:

              \begin{itemize}
                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{createTriplestore(cookie, \; triplestoreID)}$: %same as in V1

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{uploadTriplestoreContent(cookie, \; issuer, \;  triplestoreID, }$ \linebreak $\mathbf{ schema, \; syntax, \; content)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{deleteTriplestore(cookie, \; issuer, \;  triplestoreID, \; inference)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{fetchTriplestoreSchema(cookie, \;  triplestoreID, \; inference)}$:

                  \item $\{\textrm{SUCCESS}, \bot\} \leftarrow \mathbf{answerSPARQLQuery(cookie, \;  triplestoreID, \; issuer, }$\linebreak $\mathbf{ query, \; inference, \; transitivityDepth, \; expansionDepth)}$:

              \end{itemize}

        \item TRIPLESTORE protocol is a tuple with all \textrm{SEnTriS-V1.TRIPLESTORE} algorithms, apart from the $\textrm{TRIPLESTORE}.prepareSearch$. In this version the algorithm, we mask the bindings of a search. We will only define this algorithm, and assume the others to be equal as their previous definition:
              \begin{itemize}
                  \item $\{searchID, \bot\} \leftarrow \mathbf{prepareMaskedSearch(triplestoreID, \; trapdoors, \; mask, \; at)}$:
              \end{itemize}

        \item PROXY is a tuple of 2 polynomial time algorithms. Like in \gls{SEnTriS} V1 it will prepare and evaluate queries. It differs from the previous version, on query evaluation algorithm. We will only define this algorithms, and assume the others to be equal as their previous definition:
              \begin{itemize}
                  \item $\{sparqlResult, \bot\} \leftarrow \mathbf{answerSPARQLQuery(plan, K_{DGK_{eq}}, \; at)}$ takes as input a collection of a \gls{SPARQL} query plan, a secret key and an access token. If the token is valid, and grants sufficient permissions, executes the query plan and generates a query result. In case of success outputs the result, $\bot$ otherwise.
              \end{itemize}
    \end{itemize}
\end{definition}

\section{Security Analysis}

\subsection{SEnTriS Security Properties}
In conformity with the assumptions explained in ..., and based on its definition, we expect all implementations of \gls{SEnTriS} to possess the following \emph{security properties}:

\begin{enumerate}
    \item A user will be able to create triplestores and access tokens, if and only if they have an active session.
    \item A user will be able to interact with a triplestore, if and only if they have a valid access token.
    \item A user will not be able to query triplestores which they have not been given access to.
    \item A user will only be able to execute operations according to its role.
    \item An external observer will not be able to learn any useful information from eavesdropping messages sent between components.
    \item An attacker will not be able to impersonate any component.
    \item An attacker will not be able to tamper with messages sent between components.
\end{enumerate}

In conformity with the lack of data confidentiality, present in \gls{SEnTriS}-V0, the following security properties are exclusive for the other two implementations:
\begin{enumerate}
    \item An attacker with access to the \emph{Triplestore} component will learn nothing more than the size of a triplestore and its \emph{search} and \emph{access} patterns.
    \item Regarding \emph{term equality} during intermediate set operations between \emph{bindings}, an attacker with access to the \emph{Proxy} component will learn nothing more than:
        \begin{enumerate}
            \item \emph{SEnTriS V1}: the equality of all terms, belonging to a series of queries, targeting the same triplestore.
            \item \emph{SEnTriS V2} the equality of all terms present in the graph patterns of individual queries. The bindings of different queries are indistinguishable.
        \end{enumerate}
\end{enumerate}

% Plan V0 [[search_0,0, ... search0,n]->[random join_pipeline]^K -> [UNION|OPTIONAL|MINUS|JOIN search0,0, join_0,n, join,k,n]-> DISTINCT | PROJECT | ORDERBY  | SLICE]

\subsection{Security \& Performance trade-offs}

All implementations come with a trade-off between data confidentiality, query evaluation leakage and performance.

\gls{SEnTriS} V0 only provides access control, trading data confidentiality for performance.

\gls{SEnTriS} V1 provides both data confidentiality and access control, but suffers from the same leakage problems of \emph{CryptDB}, where an adversary can associate the leakage produced by multiple queries targeting the same triplestore. This facilitates the computation of the frequency distribution of all observed terms, which simplifies the execution of \emph{leakage-abuse} attacks \autocite{Jutla2021-af, Hahn2019-sg, Shafieinejad2021-rb}. This can be especially problematic during \gls{SPARQL} queries evaluation, as they tend to require a high volume of equality checks. A situation further aggravated when we apply query rewriting techniques: the modified queries usually require an even higher number of equality joins.

In \gls{SEnTriS} V2 provides both data confidentiality, access control and reduced leakage, but this comes at the cost of an extra round of communication where we have to reverse the masking applied to the \emph{equality tags} and need to fetch the values being referenced from these. Only after this step, can we decrypt the actual query results.

\subsection{Leakage Analysis}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}
\subsubsection{Discussion}
%Corolary: sentry v2 leakage is reduced

\section{Complexity Analysis}
\subsection{Storage Complexity}
%Space complexity in terms of RDF terms
\subsubsection{SEnTriS V0}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}

\subsection{Query Evaluation Complexity}
%Time complexity in terms of node search, number of joins and modular exponentiations. Also consider transitivity and expansion depth
\subsubsection{SEnTriS V0}
\subsubsection{SEnTriS V1}
\subsubsection{SEnTriS V2}

\subsection{Discussion}

\section{Summary}


